---
title: "Binary Classification of Loan Default Using Machine Learning"
author:
    - name: "storeraa17"
      affiliation: "n/a"
      email: "storera@miamioh.edu"
data: "`r format(Sys.Date(), '%B %d, %Y')`"

abstract: | 
    Abstract 
    will 
    go 
    right
    here
    and
    include
    data
    and 
    the
    purpose 
    of 
    this

keywords:
    - examples of keywords

output:
    html_document:
        toc: true
        toc_depth: 3
        toc_float: true
        number_sections: true
        theme: flatly
        highlight: tango
        code_folding: hide
        df_print: paged
        fig_width: 8
        fig_height: 5
        fig_caption: true
        self_contained: true

    pdf_document:
        toc: true
        toc_depth: 3
        number_sections: true
        latex_engine: xelatex
        fig_width: 8
        fig_height: 5
        fig_caption: true
        keep_tex: false

    word_document:
        toc: true
        toc_depth: 3
        fig_caption: true
        reference_docx: "template.docx"
 
 params:
    company_name: "loan default company name goes here"
    years_covered: "Min_Year - Max_Year"
    currency: "USD"
    include_apendix: true
    min_tdc_threshold: 1000000
    analysis_date: !r Sys.Date()

bibliography: references.bib
csl: apa.csl
link-citations: true

always_allow_html: true
editor_options:
    chunk_output_type: console
    markdown:
        wrap: 72
    
knit: "rmarkdown::render"
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE,warning=FALSE)
knitr::opts_chunk$set(fig.width=6, fig.height=4)

if(require(pacman)==0)
   {install.packages("pacman")}
pacman::p_load(devtools,caret,cluster,dplyr,fastDummies,leaps,pacman,tidyverse,skimr,fastDummies,GGally,DataExplorer,ggrepel,ggthemes,dslabs,lubridate,janitor,e1071,scales,recipes,magrittr,forcats,tidycensus,zipcodeR,tidygeocoder,tigris,sf, purrr,ROCR,rpart,rpart.plot, doParallel)

if (!require(mlba)) {
  library(devtools)
  install_github("gedeck/mlba/mlba", force=TRUE)
}
pacman::p_load(mlba,tidyverse)
```

---

# Defining the Project

## The Objective

The objective of the project is to optimize F1 score.


## Data Dictionary

| **Variable**              | **Type**  | **Description**                                                                 |
|--------------------------|----------:|:--------------------------------------------------------------------------------|
| `loan_default`           | Binary     | Whether the loan has defaulted. Values: "Yes", "No".                      |
| `term`                   | Binary     | The number of payments on the loan (36 or 60 months).          |
| `debt_settlement_flag`   | Binary     | Whether borrower is in a debt settlement program (Y/N).        |
| `home_ownership`         | Factor     | Home ownership status                                          |
| `verification_status`    | Factor     | Whether income was verified                                    |
| `purpose`                | Factor     | Category provided by the borrower for the loan request.        |
| `initial_list_status`    | Factor     | Initial listing status of the loan                             |
| `application_type`       | Factor     | Type of application                                            |
| `issue_year`             | Factor     | Number of years the loan has been issued                       |
| `revol_util`             | Numeric    | Revolving line utilization rate (credit used / total credit).  |
| `total_acc`              | Numeric    | Total number of credit lines in the borrower’s credit file.    |
| `pub_rec_bankruptcies`   | Numeric    | Number of public record bankruptcies.                          |
| `delinq_2yrs`            | Numeric    | Number of 30+ days delinquency incidents in past 2 years.      |
| `inq_last_6mths`.        | Numeric    | Number of credit inquiries in the past 6 months.               |
| `mths_since_last_delinq` | Numeric    | Months since last delinquency (NA if never).                   |
| `acc_now_delinq`         | Numeric    | Does the customer have a certificate of deposit account? (0 = No, 1 = Yes).     |
| `log_loan_amnt`          | Numeric    | The log of the annual income.                                  |
| `emp_length_numeric`     | Numeric    | Employment length in years                                    |
| `log_annual_inc`         | Numeric    | The log of the annual income.             |
| `dti_cap`                | Numeric    | Maximized dti value we capped dti values at. To take into accont of unrealistic outliers.            |
| `dti_outlier_flag`       | Numeric    | Flag to show dti outliers. To take into accont of unrealistic outliers.      |
| `open_acc_capped`        | Numeric    | Maximized open account value we capped open accounts at the log of the annual income.             |
| `open_acc_outlier_flag`  | Numeric    | The capture of the open account value being an outlier   |
| `pub_rec_capped`         | Numeric    | Number of derogatory public records with a cap.          |
| `pub_rec_outlier_flag`   | Numeric    | Whether the value captured an outlier or not.     |
| `log_revol_bal_capped`   | Numeric    | Transformed total revolving balance |
| `log_revol_bal_outlier_flag` | Numeric |  Transformed total revolving balance indicator if originally value was an outlier       |
| `mort_acc_capped`        | Numeric    | Number of mortgage accounts with a cap to handle outliers             |
| `mort_acc_outlier_flag`  | Numeric    | Indicator for whether the value was an outlier   |
| `sub_grade_score`        | Numeric    | The sub grade and grade encapsulated score.       |
| `fico_score`             | Numeric    | The average between the sum of the upper and lower ends of the borrow's FICO score range     |
| `installment_to_loanamt` | Numeric    | Ratio of installment monthly payment owed by the borrower if the loan originates to loan amount |
| `dti_annual_inc_interaction`        | Numeric    | Interaction between dti and annual income.             |
| `credit_history_years`   | Numeric    | Years between most recent credit pull date and earliest credit pull date 

## Expected Outcome

Predictions for loan default.

---

# Obtain Data

## Load the Data

```{r}
df_raw <- read_csv("train.csv", show_col_types = FALSE)
```

## Preview data

```{r}
str(df_raw)
```

## Convert Response Variable to Factor
```{r}
df_raw <- df_raw %>%
  mutate(loan_default=factor(loan_default, levels=c("No","Yes")))
```