---
title: "Binary Classification of Loan Default Using Machine Learning"
author: "storeraa17"
data: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
    The abstract should be made of
    - purpose
    - main methods
    - experimental results
    - key discussion points
    - final conclusions
keywords:
    - Predictive Statistics
    - Binary Classification
    - Machine Learning
    - Model Evaluation
    - Data Cleaning
    - Data Processing
    - Data Visualization
output:
  html_document:
    toc: true
    toc_depth: 7
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: hide
    df_print: paged
    fig_width: 8
    fig_height: 5
    fig_caption: true
    self_contained: true
  word_document:
    toc: true
    fig_caption: true
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

```{r setup, include=FALSE, message=FALSE}
```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE)
knitr::opts_chunk$set(fig_width=6,fig_height=4)

if(require(pacman)==0)
    {install.packaghes("pacman")}

pacman::p_load(
    caret,
    corrplot,
    cluster,
    DataExplorer,
    devtools,
    doParallel,
    dplyr,
    dslabs,
    e1071,
    fastDummies,
    forcats,
    GGally,
    ggplot2,
    ggthemes,
    ggrepel,
    janitor,
    leaps,
    lubridate,
    magrittr,
    mlba,
    pacman,
    recipes,
    rpart,
    rpart.plot,
    ROCR,
    purrr,
    scales,
    sf,
    skimr,
    tidycensus,
    tidygeocoder,
    tidyverse,
    tigris,
    zipcodeR
    # Add more packages as needed here
)

set.seed(123)

project_dir <- "ML_portfolio/R/loan_default/data"
```

---

# **Introduction**

The purpose of this project is to support better loan risk assessment. A reliable default prediction model can help lenders make more informed decisions, reduce financial losses, and improve consistency in loan evaluation. Rather than relying only on manual review or a limited set of business rules, a predictive model can evaluate many borrower and loan characteristics at the same time. Thus, the baility to predict loan default is a critical problem. For example, financial analysts may need to estimate the risk that a borrow will fail to repay a loan. When a loan is issued, the lender accepts financial risk based on information about the borrower, the loan terms, credit history, income, debt obligations, and application characteristics. If risk is underestimated, the lender may approve loans that are more likely to default. If risk is overestimated, the lender may reject borrowers who could have successfully repaid their loans.  Herein, we report a machine learning model used for predictive binary classification of loan default.

The machine learning model used to evaluated will be subjected to several inputs to supply the desired loan default classification output. The make-up inputs cover several important features to consider, such as borrow information, loan terms, finances, credit history, accounting, delinquencies, administration, and status variables. The data representing these variables is split into two datasets:

1. `train.csv`
2. 'holdout.csv`

The dataset 'train.csv` represents the training data and will be the dataset used for model learning. While, `holdout.csv` will be the dataset that the model is evaluated on. The evaluation of the `holdout.csv` dataset will be done on the premise of F1 score primarily. However, F1 score is not exclusively the only metric used to determine model performance. Secondary evaluation metrics used to examine model performance include accuracy, balanced accuracy, recall/sensitivity, precision, number of false positives, number of false negatives, and AUC of an ROC curve.

These performance metrics will be used to evaluate decision trees, tree ensembles, logistic regression, and neural network models. The constructions of these models can be for in the `model` folder. Nevertheless, the best resulting models, based on F1 score, will be imported and model performances will be compared to determine the optimal model.

The variables found in `train.csv` and `holdout.csv` are described in the data dictionary below:

| Variable | Type | Description |
|:---|:---|:---|
| `loan_default` | Categorical | Indicates whether the borrower defaulted. Values are `Yes` and `No`. |
| `emp_title` | Categorical / text | Borrower's reported job title. High-cardinality text variable. |
| `emp_length` | Categorical / ordinal | Borrower's reported employment length. |
| `home_ownership` | Categorical | Borrower's home ownership status. |
| `annual_inc` | Numeric | Borrower's self-reported annual income. |
| `verification_status` | Categorical | Whether income or financial information was verified. |
| `address` | Categorical / text | Borrower's recorded address. High-cardinality text variable. |
| `loan_amnt` | Numeric | Original loan principal amount. |
| `term` | Categorical | Scheduled repayment length, generally 36 or 60 months. |
| `int_rate` | Numeric | Annual interest rate assigned to the loan. |
| `installment` | Numeric | Scheduled monthly payment. |
| `grade` | Categorical / ordinal | Broad lender-assigned credit grade. |
| `sub_grade` | Categorical / ordinal | More detailed credit grade. |
| `purpose` | Categorical | Borrower's stated loan purpose. |
| `title` | Categorical / text | Borrower-provided title for the loan request. |
| `dti` | Numeric | Debt-to-income ratio. |
| `earliest_cr_line` | Date-like character | Date of borrower's earliest reported credit line. |
| `fico_range_low` | Numeric | Lower bound of reported FICO range. |
| `fico_range_high` | Numeric | Upper bound of reported FICO range. |
| `inq_last_6mths` | Numeric | Number of credit inquiries in the last six months. |
| `mths_since_last_delinq` | Numeric | Months since most recent delinquency. Missing values may indicate no prior delinquency or unavailable information. |
| `last_credit_pull_d` | Date-like character | Most recent credit pull date. |
| `open_acc` | Numeric | Number of open credit accounts. |
| `pub_rec` | Numeric | Number of derogatory public records. |
| `revol_bal` | Numeric | Total revolving balance. |
| `revol_util` | Numeric | Revolving credit utilization rate. |
| `total_acc` | Numeric | Total number of credit accounts. |
| `mort_acc` | Numeric | Number of mortgage accounts. |
| `pub_rec_bankruptcies` | Numeric | Number of public-record bankruptcies. |
| `delinq_2yrs` | Numeric | Number of delinquency incidents in the last two years. |
| `acc_now_delinq` | Numeric | Number of accounts currently delinquent. |
| `issue_d` | Date-like character | Month and year the loan was issued. |
| `initial_list_status` | Categorical | Initial listing status assigned to the loan. |
| `application_type` | Categorical | Individual, joint, or direct-pay application type. |
| `hardship_flag` | Categorical | Indicates hardship arrangement status. |
| `debt_settlement_flag` | Categorical | Indicates debt-settlement status. |

---

# **Methods**

## *Data Loading*

The datasets, `train.csv` and `holdout.csv`, were loaded and the target variable. `loan_default`, was immediately converted into a factor.

```{r data loading}
train_raw <- readr::read_csv(
  file.path(project_dir, "train.csv"),
  show_col_types = FALSE) %>%
  mutate(loan_default = factor(loan_default, levels = c("No", "Yes")))
```

```{r data loading}
hold_raw <- readr::read_csv(
  file.path(project_dir, "holdout.csv"),
  show_col_types = FALSE) %>%
  mutate(loan_default = factor(loan_default, levels = c("No", "Yes")))
```

## *Data Inspection*

### Previewing the Dataset
____________________________________________________________________________________________________________

The structure of the datasets were assessed using the `str()` function. Subsequent previewing of the datasets was done through the utilization of the `head()` function, which shows the first few rows of a dataset.

```{r data inspection}
str(train_raw)
```

```{r data inspection}
str(hold_raw)
```

```{r data inspection}
head(train_raw)
```

```{r data inspection}
head(hold_raw)
```

### Dataset Dimensions
____________________________________________________________________________________________________________

The dimensions of the dataset were determined by employed the `dim()` function.

```{r data inspection}
dim(train_raw)
```

```{r data inspection}
dim(hold_raw)
```

### The Number of Unique Instances
____________________________________________________________________________________________________________
The number of unique instances for each variable were examined via `sapply()` and `n_distinct()` functions.

```{r data inspection}
sapply(train_raw, n_distinct)
```

```{r data inspection}
sapply(hold_raw, n_distinct)
```

### Class Comparison of Variables
____________________________________________________________________________________________________________
It is important to compare the classes of each variable in the training and holdout datasets to ensure consistency. This can be done using the `sapply()` and `class()` functions. To present the findings more cleanly, a tibble can be created to display the variable names, their corresponding classes in the training dataset, and their classes in the holdout dataset. The function `tibble::tibble()` can be used to create this tibble, and the `sapply()` function can be used to apply the `class()` function to each variable in both datasets. The resulting tibble will provide a clear comparison of the variable classes between the two datasets.

```{r data inspection}
class_comparison <- tibble::tibble(
  Variable = names(train_raw),
  Train_Class = sapply(train_raw, function(x) class(x)[1]),
  Holdout_Class = sapply(hold_raw, function(x) class(x)[1])
)

class_comparison
```

## *Summary Statistics*

### Summary Statistics
____________________________________________________________________________________________________________

The summary statistics of the datasets were examined using the `skimr::skim()` function. This function provides a comprehensive overview of the data, including the number of missing values, mean, median, standard deviation, and other relevant statistics for each variable.

```{r data }
skimr::skim(train_raw)
```

```{r data processing}
skimr::skim(hold_raw)
```

## *Data Visualization**

### Grouping Numeric Variables
____________________________________________________________________________________________________________

The numeric variables in the training dataset were identified and grouped together for further analysis. This was done using the `select()` function from the `dplyr` package, along with the `where()` function to filter for numeric variables. The names of these numeric variables were then extracted using the `names()` function.

```{r data visualization}
numeric_vars <- train_raw %>%
  select(where(is.numeric)) %>%
  names()

numeric_vars
```

The numeric variables in the holdout dataset were also identified and grouped together for further analysis. This was done using the `select()` function from the `dplyr` package, along with the `where()` function to filter for numeric variables. The names of these numeric variables were then extracted using the `names()` function.

```{r data visualization}
hold_numeric_vars <- hold_raw %>%
  select(where(is.numeric)) %>%
  names()

hold_numeric_vars
```

### Grouping Categorical Variables
____________________________________________________________________________________________________________

The categorical variables in the training dataset were identified and grouped together for further analysis. This was done using the `select()` function from the `dplyr` package, along with the `where()` function to filter for character or factor variables. The names of these categorical variables were then extracted using the `names()` function.

```{r data visualization}
categorical_vars <- train_raw %>%
  select(where(~ is.character(.x) || is.factor(.x))) %>%
  names()

categorical_vars
```

The categorical variables in the holdout dataset were also identified and grouped together for further analysis. This was done using the `select()` function from the `dplyr` package, along with the `where()` function to filter for character or factor variables. The names of these categorical variables were then extracted using the `names()` function.

```{r data visualization}
hold_categorical_vars <- hold_raw %>%
  select(where(~ is.character(.x) || is.factor(.x))) %>%
  names()

hold_categorical_vars
```

### Target Variable Distribution
____________________________________________________________________________________________________________

#### Distribution of the Target Variable in the Training Dataset

The distribution of the target variable, `loan_default`, in the training dataset was examined using the `table()` function to count the occurrences of each class, and the `prop.table()` function to calculate the proportions. The results were then visualized using a bar plot created with the `ggplot2` package.

```{r data visualization}
train_target_distribution <- table(train_raw$loan_default)
train_target_proportions <- prop.table(train_target_distribution)
```

```{r data visualization}
train_target_distribution_df <- as.data.frame(train_target_distribution)
colnames(train_target_distribution_df) <- c("loan_default", "Count")
```

```{r data visualization}
ggplot(train_target_distribution_df, aes(x = loan_default, y = Count, fill = loan_default)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = c("No" = "#1f77b4", "Yes" = "#ff7f0e")) +
    labs(title = "Distribution of Loan Default in Training Dataset",
        x = "Loan Default",
        y = "Count") +
    theme_bw(base_size = 14) +
    theme(legend.position = "none",
          plot.title = element_text(face = "bold", hjust = 0.5))
```

#### Distribution of the Target Variable in the Holdout Dataset

The distribution of the target variable, `loan_default`, in the holdout dataset was examined using the `table()` function to count the occurrences of each class, and the `prop.table()` function to calculate the proportions. The results were then visualized using a bar plot created with the `ggplot2` package.

```{r data visualization}
hold_target_distribution <- table(hold_raw$loan_default)
hold_target_proportions <- prop.table(hold_target_distribution)
```

```{r data visualization}
hold_target_distribution_df <- as.data.frame(hold_target_distribution)
colnames(hold_target_distribution_df) <- c("loan_default", "Count")
```

```{r data visualization}
ggplot(hold_target_distribution_df, aes(x = loan_default, y = Count, fill = loan_default)) +
    geom_bar(stat = "identity", color="white", alpha=0.9) +
    scale_fill_manual(values = c("No" = "#1f77b4", "Yes" = "#ff7f0e")) +
    labs(title = "Distribution of Loan Default in Holdout Dataset",
        x = "Loan Default",
        y = "Count") +
    theme_bw(base_size = 14) +
    theme(legend.position = "none",
          plot.title = element_text(face = "bold", hjust = 0.5))
```

### Numeric Variable Distributions - Histograms
____________________________________________________________________________________________________________

The numeric variables in the training and holdout datasets were examined using histograms to visualize their distributions. The `DataExplorer` package was used to create the histograms, with customizations applied to the theme and appearance of the plots.

#### Historgrams of Numeric Variables in the Training Dataset

```{r data visualization}
DataExplorer::plot_histogram(
  train_raw %>% select(all_of(numeric_vars)),
  ggtheme =
    theme_bw(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "#1F77B4"),
      strip.text = element_text(color = "white", face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey90")
    ),
  geom_histogram_args = list(
    fill = "#4E79A7",
    color = "white",
    alpha = 0.9
  ),
  ncol = 3
)
```

#### Historgrams of Numeric Variables in the Holdout Dataset

```{r data visualization}
DataExplorer::plot_histogram(
  hold_raw %>% select(all_of(hold_numeric_vars)),
  ggtheme =
    theme_bw(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "#a3231e"),
      strip.text = element_text(color = "white", face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey90")
    ),
  geom_histogram_args = list(
    fill = "#a3231e",
    color = "white",
    alpha = 0.9
  ),
  ncol = 3
)
```

### Numeric Variable Distributions - Boxplots
____________________________________________________________________________________________________________

The numeric variables in the training and holdout datasets were examined using boxplots to visualize their distributions and identify potential outliers. The `DataExplorer` package was used to create the boxplots, with customizations applied to the theme and appearance of the plots.

#### Boxplots of Numeric Variables in the Training Dataset

```{r data visualization}
DataExplorer::plot_boxplot(
  train_raw %>% 
    select(loan_default, all_of(numeric_vars)),
  by = "loan_default",
  ggtheme =
    theme_bw(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "#1F77B4"),
      strip.text = element_text(color = "white", face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey90")
    ),
  geom_boxplot_args = list(
    fill = "#4E79A7",
    color = "white",
    alpha = 0.9
  ),
  ncol = 3
)
```

#### Boxplots of Numeric Variables in the Holdout Dataset

```{r data visualization}
DataExplorer::plot_boxplot(
  hold_raw %>% 
    select(loan_default, all_of(hold_numeric_vars)),
  by = "loan_default",
  ggtheme =
    theme_bw(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "#a3231e"),
      strip.text = element_text(color = "white", face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey90")
    ),
  geom_boxplot_args = list(
    fill = "#a3231e",
    color = "white",
    alpha = 0.9
  ),
  ncol = 3
)
```

### Numeric Variable Distributions - Density Plots
____________________________________________________________________________________________________________

The numeric variables in the training and holdout datasets were examined using density plots to visualize their distributions. The `DataExplorer` package was used to create the density plots, with customizations applied to the theme and appearance of the plots.

#### Density Plots of Numeric Variables in the Training Dataset

```{r data visualization}
DataExplorer::plot_density(
  train_raw %>% select(all_of(numeric_vars)),
  ggtheme =
    theme_bw(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "#1F77B4"),
      strip.text = element_text(color = "white", face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey90")
    ),
  geom_density_args = list(
    fill = "#4E79A7",
    color = "white",
    alpha = 0.9
  ),
  ncol = 3
)
```

#### Density Plots of Numeric Variables in the Holdout Dataset

```{r data visualization}
DataExplorer::plot_density(
  hold_raw %>% select(all_of(hold_numeric_vars)),
  ggtheme =
    theme_bw(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "#a3231e"),
      strip.text = element_text(color = "white", face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey90")
    ),
  geom_density_args = list(
    fill = "#a3231e",
    color = "white",
    alpha = 0.9
  ),
  ncol = 3
)
```

### Numeric Variable Distributions - Heat Map
____________________________________________________________________________________________________________

The numeric variables in the training and holdout datasets were examined using correlation heat maps to visualize the relationships between the variables. The `DataExplorer` package was used to create the heat maps, with customizations applied to the theme and appearance of the plots.

#### Correlation Heat Map of Numeric Variables in the Training Dataset

```{r data visualization}
DataExplorer::plot_correlation(
  train_raw %>% select(all_of(numeric_vars)),
  ggtheme =
    theme_bw(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "#1F77B4"),
      strip.text = element_text(color = "white", face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey90")
    ),
  cor_args = list(
    use = "pairwise.complete.obs",
    method = "pearson"
  )
)
```

#### Correlation Heat Map of Numeric Variables in the Holdout Dataset

```{r data visualization}
DataExplorer::plot_correlation(
  hold_raw %>% select(all_of(hold_numeric_vars)),
  ggtheme =
    theme_bw(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "#a3231e"),
      strip.text = element_text(color = "white", face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey90")
    ),
  cor_args = list(
    use = "pairwise.complete.obs",
    method = "pearson"
  )
)
```

### Numeric Variable Distributions - Correlation Pairs
____________________________________________________________________________________________________________

The numeric variables in the training and holdout datasets were examined using correlation matrices to visualize the relationships between the variables. The `cor()` function was used to calculate the correlation coefficients, with the `use` argument set to "pairwise.complete.obs" to handle missing values. The resulting correlation matrices were then visualized using heat maps created with the `corrplot` package.

#### Correlation Matrix of Numeric Variables in the Training Dataset

```{r data visualization}
correlation_matrix_train <- cor(train_raw %>% select(all_of(numeric_vars)), use = "pairwise.complete.obs")
```

##### Correlation Matrix of Numeric Variables in the Holdout Dataset
```{r data visualization}
correlation_matrix_hold <- cor(hold_raw %>% select(all_of(hold_numeric_vars)), use = "pairwise.complete.obs")
```

### Categorical Variable Distributions - Frequency Tables
____________________________________________________________________________________________________________

The categorical variables in the training and holdout datasets were examined using frequency tables to visualize their distributions. The `table()` function was used to create the frequency tables, with customizations applied to the theme and appearance of the tables.

#### Frequecy Tables of Categorical Variables in the Training Dataset

```{r data visualization}
for (col in names(categorical_vars)) {
    print(table(categorical_vars[[col]])[1:5])
}
```

#### Frequecy Tables of Categorical Variables in the Holdout Dataset

```{r data visualization}
for (col in names(hold_categorical_vars)) {
    print(table(hold_categorical_vars[[col]])[1:5])
}
```

### Categorical Variable Distributions - Bar Plots
____________________________________________________________________________________________________________

The categorical variables in the training and holdout datasets were examined using bar plots to visualize their distributions. The `ggplot2` package was used to create the bar plots, with customizations applied to the theme and appearance of the plots. A function was created to generate bar plots for each categorical variable in the datasets. The function takes three arguments: the dataset, the list of categorical variables, and a title prefix for the plots. It iterates through each categorical variable, creates a bar plot using `ggplot2`, and prints the plot.

#### Creating the Bar Plots Function

```{r data visualization}
bar_plots <- function(data, categorical_vars, title_prefix, max_levels = 20) {
  for (col in categorical_vars) {
    
    n_levels <- data %>%
      distinct(.data[[col]]) %>%
      nrow()
    
    if (n_levels > max_levels) {
      message("Skipping ", col, ": ", n_levels, " unique levels.")
      next
    }
    
    plot <- ggplot(data, aes(x = .data[[col]])) +
      geom_bar(fill = "#4E79A7", color = "white", alpha = 0.9) +
      labs(
        title = paste(title_prefix, col),
        x = col,
        y = "Count"
      ) +
      theme_bw(base_size = 14) +
      theme(
        legend.position = "none",
        plot.title = element_text(face = "bold", hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1)
      )

    print(plot)
  }
}
```

#### Bar Plots of Categorical Variables in the Training Dataset

```{r data visualization}
bar_plots(
  train_raw,
  categorical_vars,
  "Distribution of Categorical Variable in Training Dataset:"
)
```

#### Bar Plots of Categorical Variables in the Holdout Dataset

```{r data visualization}
bar_plots(
  hold_raw,
  hold_categorical_vars,
  "Distribution of Categorical Variable in Holdout Dataset:"
)
```

### Categorial Variable Distributions - Proportions by Loan Default Status
____________________________________________________________________________________________________________

#### Proportions of Categorical Variables by Loan Default Status in the Training Dataset

```{r data visualization}
for (col in names(categorical_vars)) {
    prop_table <- prop.table(table(train_raw[[col]], train_raw$loan_default), margin = 1)
    print(paste("Proportions of", col, "by Loan Default Status in Training Dataset:"))
    print(prop_table)
}
```

#### Proportions of Categorical Variables by Loan Default Status in the Holdout Dataset

```{r data visualization}
for (col in names(hold_categorical_vars)) {
    prop_table <- prop.table(table(hold_raw[[col]], hold_raw$loan_default), margin = 1)
    print(paste("Proportions of", col, "by Loan Default Status in Holdout Dataset:"))
    print(prop_table)
}
```

## *Data Quality Assessment*

### Missing Value Counts
____________________________________________________________________________________________________________

The presence of missing values in the training and holdout datasets was examined using the `sapply()` function to apply the `n_missing()` function to each variable. The `n_missing()` function counts the number of missing values in a variable. The results were then presented in a table format for easy interpretation.

#### Missing Values in the Training Dataset

```{r dqa}
sapply(train_raw, n_missing)
```

#### Missing Values in the Holdout Dataset

```{r dqa}
sapply(hold_raw, n_missing)
```

### Missing Value Plots
____________________________________________________________________________________________________________

The missing values in the training and holdout datasets were visualized using the `plot_missing()` function from the `DataExplorer` package. This function creates a plot that shows the proportion of missing values for each variable in the dataset. The plots were customized using the `ggtheme` argument to apply a specific theme and appearance. The percentage of missing values for each variable is displayed on the y-axis, while the variable names are displayed on the x-axis. The plots provide a visual representation of the extent of missing data in the datasets, which can inform decisions regarding data cleaning and preprocessing.

#### Missing Value Plots in the Training Dataset

```{r dqa}
DataExplorer::plot_missing(train_raw, ggtheme = theme_bw(base_size = 14))
```

#### Missing Value Plots in the Holdout Dataset

```{r dqa}
DataExplorer::plot_missing(hold_raw, ggtheme = theme_bw(base_size = 14))
```

### Anomalic Values Assessment
____________________________________________________________________________________________________________

The strategy employed to detect anomalic values in the numeric variables of the training and holdout datasets was based on the interquartile range (IQR) method. The IQR is a measure of statistical dispersion, which is the difference between the third quartile (Q3) and the first quartile (Q1). Anomalic values, or outliers, were identified as any data points that fall below Q1 - 1.5 * IQR or above Q3 + 1.5 * IQR. The `boxplot.stats()` function was used to calculate the IQR and identify outliers for each numeric variable in both datasets. The results were then printed for each variable that contained anomalic values. This only applies to numeric variables, as categorical variables do not have a meaningful IQR. Outliers are expected in financial data. High income, high revolving balance, and large loan amounts can be valid borrower records rather than data entry errors. For this reason, the workflow avoids deleting outliers by default. The model search includes variants that add outlier flags and log transformations.

#### Anomalic Values in the Training Dataset

```{r dqa}
for (col in names(numeric_vars)) {
    outliers <- boxplot.stats(train_raw[[col]])$out
    if (length(outliers) > 0) {
        print(paste("Anomalic values in", col, "in Training Dataset:"))
        print(outliers)
    }
}
```

#### Anomalic Values in the Holdout Dataset

```{r dqa}
for (col in names(hold_numeric_vars)) {
    outliers <- boxplot.stats(hold_raw[[col]])$out
    if (length(outliers) > 0) {
        print(paste("Anomalic values in", col, "in Holdout Dataset:"))
        print(outliers)
    }
}
```

### Inconsistent Values Assessment
_____________________________________________________________________________________________________________

First, the categorical variables in both the training and holdout datasets were checked for inconsistent values. Inconsistent values refer to unexpected or invalid entries in categorical variables that do not conform to the expected categories. The `unique()` function was used to identify the unique values present in each categorical variable. If a variable had more than 10 unique values, it was flagged for further inspection, as this may indicate potential inconsistencies or data quality issues. The results were printed for each variable that contained inconsistent values. Second, the numeric variables in both datasets were checked for inconsistent values. Inconsistent values in numeric variables refer to unexpected or invalid entries that fall outside the expected range or distribution of values. The `summary()` function was used to examine the summary statistics of each numeric variable, including the minimum, maximum, mean, and quartiles. If a variable had extreme values or outliers that were significantly different from the rest of the data, it was flagged for further inspection. The results were printed for each variable that contained inconsistent values. This step is important to ensure the integrity and reliability of the data before proceeding with further analysis and modeling.

#### Inconsistent Values in Categorical Variables in the Training Dataset

```{r dqa}
for (col in names(categorical_vars)) {
    unique_values <- unique(train_raw[[col]])
    if (length(unique_values) > 10) {
        print(paste("Inconsistent values in", col, "in Training Dataset:"))
        print(unique_values)
    }
}
```

#### Inconsistent Values in Categorical Variables in the Holdout Dataset

```{r dqa}
for (col in names(hold_categorical_vars)) {
    unique_values <- unique(hold_raw[[col]])
    if (length(unique_values) > 10) {
        print(paste("Inconsistent values in", col, "in Holdout Dataset:"))
        print(unique_values)
    }
}
```

#### Inconsistent Values in Numeric Variables in the Training Dataset

```{r dqa}
for (col in names(numeric_vars)) {
    summary_stats <- summary(train_raw[[col]])
    if (summary_stats["Min."] < 0 || summary_stats["Max."] > 1000000) {
        print(paste("Inconsistent values in", col, "in Training Dataset:"))
        print(summary_stats)
    }
}
```

#### Inconsistent Values in Numeric Variables in the Holdout Dataset

```{r dqa}
for (col in names(hold_numeric_vars)) {
    summary_stats <- summary(hold_raw[[col]])
    if (summary_stats["Min."] < 0 || summary_stats["Max."] > 1000000) {
        print(paste("Inconsistent values in", col, "in Holdout Dataset:"))
        print(summary_stats)
    }
```

### High Cardinality Assessment
_____________________________________________________________________________________________________________

A categorical variable with a high cardinality is one that has a large number of unique categories or levels. High cardinality can pose challenges for modeling, as it may lead to overfitting or increased computational complexity. In this analysis, categorical variables with more than 10 unique values were flagged as high cardinality. The `unique()` function was used to identify the unique values present in each categorical variable, and the results were printed for each variable that had high cardinality. These variables can be further examined to determine if they can be transformed or encoded in a way that reduces their cardinality, such as through grouping or feature engineering. The presence of high cardinality categorical variables during modeling can be addressed through techniques such as one-hot encoding, target encoding, or embedding representations, depending on the modeling approach and the specific characteristics of the data. The high cardinality variables can limit the choice of models and may require additional preprocessing steps to ensure that the model can effectively learn from the data.

#### High Cardinality Categorical Variables in the Training Dataset

```{r dqa}
for (col in names(categorical_vars)) {
    unique_values <- unique(train_raw[[col]])
    if (length(unique_values) > 10) {
        print(paste("High cardinality categorical variable in Training Dataset:", col))
        print(unique_values)
    }
}
```

#### High Cardinality Categorical Variables in the Holdout Dataset

```{r dqa}
for (col in names(hold_categorical_vars)) {
    unique_values <- unique(hold_raw[[col]])
    if (length(unique_values) > 10) {
        print(paste("High cardinality categorical variable in Holdout Dataset:", col))
        print(unique_values)
    }
}
```

### Rare Category Assessment
_____________________________________________________________________________________________________________

The presence of rare categories can be assessed with code below. A function, `assess_rare_categories`, was created. This function has five dependencies in order to properly evaluate the presence of rare categories.

#### Creating A Rare Category Assessment Function

```{r dqa}
assess_rare_categories <- function(data,
                                   variables = NULL,
                                   target = "loan_default",
                                   min_count = 30,
                                   min_prop = 0.01) {
  if (is.null(variables)) {
    variables <- names(data)[sapply(data, function(x) is.character(x) || is.factor(x))]
  }

  variables <- setdiff(variables, target)

  purrr::map_dfr(variables, function(v) {
    data %>%
      transmute(
        variable = v,
        value = as.character(.data[[v]])
      ) %>%
      mutate(
        value = ifelse(is.na(value) | stringr::str_squish(value) == "", "Missing", value),
        value = stringr::str_squish(value)
      ) %>%
      count(variable, value, name = "count") %>%
      mutate(
        total_rows = nrow(data),
        proportion = count / total_rows,
        rare_category = count < min_count | proportion < min_prop
      )
  })
}
```

#### Rare Category Assessment of Categorical Variables in the Training Dataset

```{r dqa}
train_rare_categories <- assess_rare_categories(
  data = train_raw,
  target = "loan_default",
  min_count = 30,
  min_prop = 0.01
)

train_rare_summary <- train_rare_categories %>%
  group_by(variable) %>%
  summarise(
    total_levels = n(),
    rare_levels = sum(rare_category),
    percent_levels_rare = rare_levels / total_levels,
    rows_in_rare_levels = sum(count[rare_category]),
    percent_rows_in_rare_levels = rows_in_rare_levels / first(total_rows),
    .groups = "drop"
  ) %>%
  arrange(desc(percent_rows_in_rare_levels), desc(rare_levels))

train_rare_summary %>%
  mutate(
    percent_levels_rare = scales::percent(percent_levels_rare, accuracy = 0.1),
    percent_rows_in_rare_levels = scales::percent(percent_rows_in_rare_levels, accuracy = 0.1)
  ) %>%
  knitr::kable(caption = "Rare Category Assessment for Training Data")
```

#### Rare Category Assessment of Categorical Variables in the Holdout Dataset

```{r dqa}
holdout_rare_categories <- assess_rare_categories(
  data = hold_raw,
  target = "loan_default",
  min_count = 30,
  min_prop = 0.01
)

holdout_rare_summary <- holdout_rare_categories %>%
  group_by(variable) %>%
  summarise(
    total_levels = n(),
    rare_levels = sum(rare_category),
    percent_levels_rare = rare_levels / total_levels,
    rows_in_rare_levels = sum(count[rare_category]),
    percent_rows_in_rare_levels = rows_in_rare_levels / first(total_rows),
    .groups = "drop"
  ) %>%
  arrange(desc(percent_rows_in_rare_levels), desc(rare_levels))

holdout_rare_summary %>%
  mutate(
    percent_levels_rare = scales::percent(percent_levels_rare, accuracy = 0.1),
    percent_rows_in_rare_levels = scales::percent(percent_rows_in_rare_levels, accuracy = 0.1)
  ) %>%
  knitr::kable(caption = "Rare Category Assessment for Holdout Data")
```

### Data Leakage Assessment
_____________________________________________________________________________________________________________

It was assumed that data leakage was not a potential problem. However, this will not be the case for real-world examples. Both the training and holdout datasets were examined for possible data leakages.

#### Create Data Leakage Assessment Function

```{r dqa}
assess_data_leakage <- function(data,
                                target = "loan_default",
                                positive_class = "Yes",
                                negative_class = "No") {
  leakage_name_patterns <- c(
    "settlement", "hardship", "recover", "collection", "last",
    "payment", "pymnt", "status", "default", "chargeoff",
    "out_prncp", "total_rec", "funded"
  )

  predictors <- setdiff(names(data), target)

  name_flags <- tibble::tibble(
    variable = predictors,
    leakage_name_flag = stringr::str_detect(
      stringr::str_to_lower(variable),
      paste(leakage_name_patterns, collapse = "|")
    )
  )

  categorical_results <- purrr::map_dfr(predictors, function(v) {
    if (!(is.character(data[[v]]) || is.factor(data[[v]]))) return(NULL)

    temp <- data %>%
      mutate(
        target_value = .data[[target]] == positive_class,
        predictor_value = as.character(.data[[v]]),
        predictor_value = ifelse(is.na(predictor_value), "Missing", predictor_value)
      ) %>%
      group_by(predictor_value) %>%
      summarise(
        n = n(),
        default_rate = mean(target_value, na.rm = TRUE),
        .groups = "drop"
      )

    tibble::tibble(
      variable = v,
      variable_type = "categorical",
      max_default_rate = max(temp$default_rate, na.rm = TRUE),
      min_default_rate = min(temp$default_rate, na.rm = TRUE),
      default_rate_spread = max_default_rate - min_default_rate,
      levels = nrow(temp),
      suspicious_strength_flag = default_rate_spread >= 0.75
    )
  })

  numeric_results <- purrr::map_dfr(predictors, function(v) {
    if (!is.numeric(data[[v]])) return(NULL)

    temp <- data %>%
      select(all_of(c(target, v))) %>%
      filter(!is.na(.data[[target]]), !is.na(.data[[v]]))

    if (nrow(temp) == 0 || length(unique(temp[[v]])) < 2) return(NULL)

    roc_obj <- tryCatch(
      pROC::roc(
        response = temp[[target]],
        predictor = temp[[v]],
        levels = c(negative_class, positive_class),
        direction = "<",
        quiet = TRUE
      ),
      error = function(e) NULL
    )

    if (is.null(roc_obj)) return(NULL)

    auc_value <- as.numeric(pROC::auc(roc_obj))
    auc_value <- max(auc_value, 1 - auc_value)

    tibble::tibble(
      variable = v,
      variable_type = "numeric",
      auc = auc_value,
      suspicious_strength_flag = auc_value >= 0.90
    )
  })

  bind_rows(categorical_results, numeric_results) %>%
    left_join(name_flags, by = "variable") %>%
    mutate(
      possible_leakage_flag = leakage_name_flag | suspicious_strength_flag
    ) %>%
    arrange(desc(possible_leakage_flag), desc(suspicious_strength_flag))
}
```

#### Data Laakage Assessment of Training Dataset

```{r dqa}
train_leakage_assessment <- assess_data_leakage(
  data = train_raw,
  target = "loan_default",
  positive_class = "Yes",
  negative_class = "No"
)

train_leakage_assessment %>%
  filter(possible_leakage_flag) %>%
  mutate(
    default_rate_spread = ifelse(
      is.na(default_rate_spread),
      NA,
      scales::percent(default_rate_spread, accuracy = 0.1)
    ),
    auc = round(auc, 3)
  ) %>%
  knitr::kable(caption = "Potential Data Leakage Flags in Training Data")
```

#### Data Leakage Assessment of Holdout Dataset

```{r dqa}
holdout_leakage_assessment <- assess_data_leakage(
  data = hold_raw,
  target = "loan_default",
  positive_class = "Yes",
  negative_class = "No"
)

holdout_leakage_assessment %>%
  filter(possible_leakage_flag) %>%
  mutate(
    default_rate_spread = ifelse(
      is.na(default_rate_spread),
      NA,
      scales::percent(default_rate_spread, accuracy = 0.1)
    ),
    auc = round(auc, 3)
  ) %>%
  knitr::kable(caption = "Potential Data Leakage Flags in Holdout Data")
```

## *Data Preprocessing*

### Character-to-Factor Conversion
_____________________________________________________________________________________________________________

Categorical fields such as `term`, `emp_length`, `home_ownership`, `verification_status`, `purpose`, `initial_list_status`, and `application_type` are treated as categorical predictors. Some categorical fields are also converted into ordered or numeric scores when their ordering has business meaning.

#### Term

```{r data preprocessing}
train_raw <- train_raw %>%
  mutate(
    term = factor(term, leveels = c("36 months", "60 months"))
  )
```

#### Employee Length

```{r data preprocessing}
train_raw <- train_raw %>%
  mutate(
    emp_length = forcats::fct_na_value_to_level(factor(emp_length), level = "Missing"),
    emp_length_numeric = case_when(
        emp_length == "< 1 year" ~ 0,
        emp_length == "1 year" ~ 1,
        emp_length == "2 years" ~ 2,
        emp_length == "3 years" ~ 3,
        emp_length == "4 years" ~ 4,
        emp_length == "5 years" ~ 5,
        emp_length == "6 years" ~ 6,
        emp_length == "7 years" ~ 7,
        emp_length == "8 years" ~ 8,
        emp_length == "9 years" ~ 9,
        emp_length == "10+ years" ~ 10,
        TRUE ~ NA_real_
      )
  )
```

#### Home Ownership

```{r data preprocessing}
train_raw$home_ownership <- as.factor(train_raw$home_ownership)
```

#### Verification Status

```{r data preprocessing}
train_raw$verification_status <- as.factor(train_raw$verification_status)
```

#### Purpose

```{r data preprocessing}
train_raw$purpose <- as.factor(train_raw$purpose)
```

#### Initial List Status

```{r data preprocessing}
train_raw$initial_list_status <- as.factor(train_raw$initial_list_status)
```

#### Application Type

```{r data preprocessing}
train_raw$application_type <- as.factor(train_raw$application_type)
```

### Character-to-Numeric Conversion
_____________________________________________________________________________________________________________

#### Term Number

```{r data preprocessing}
train_raw <- train_raw %>%
  mutate(
    term_num = readr::parse_number(as.character(term))
  )
```

#### Employee Length Number

```{r data preprocessing}
train_raw <- train_raw %>%
  mutate(
    emp_length_numeric = readr::parse_number(as.character(emp_length))
  )
```

### Character-to-Date Conversion
_____________________________________________________________________________________________________________

The variables `issue_d` and `earliest_cr_line` were parsed using `lubridate`::`my()`. These parsed dates support credit history and loan timing features.

#### Issue Date

```{r data preprocessing}
train_raw <- train_raw %>%
  mutate(
    issue_d = lubridate::my(issue_d)
  )
```

#### Earliest Credit Line

```{r data preprocessing}
train_raw <- train_raw %>%
  mutate(
    earliest_cr_line = lubridate::my(earliest_cr_line)
  )
```

### Missing Value Treatment
_____________________________________________________________________________________________________________

Median imputation was employed as the imputation strategy to handle missing values. Later, this will be done on the holdout dataset. The training data median will be used on the holdout dataset as is standard protocol. 

```{r data preprocessing}
num_cols_train <- names(train_raw)[sapply(train_raw, is.numeric)]

train_medians <- sapply(train_raw[num_cols_train], median, na.rm = TRUE)

for (v in num_cols_train) {
  train_raw[[v]][!is.finite(train_raw[[v]])] <- NA_real_
  train_raw[[v]][is.na(train_raw[[v]])] <- train_medians[[v]]
```

### Handling Rare or Sparse Categories
_____________________________________________________________________________________________________________

#### Application Type

```{r data preprocessing}
train_raw <- train_raw %>%
  mutate(
    application_type = forcats::fct_collapse(
      factor(application_type),
      OTHER = c("JOINT", "DIRECT_PAY")
    )
  )
```

#### Home Ownership

```{r data preprocessing}
train_raw <- train_raw %>%
  mutate(
    home_ownership = forcats::fct_collapse(
      factor(home_ownership),
      OTHER = c("OTHER", "ANY", "NONE")
    )
  )
```

### Removing Unnecessary Variables
_____________________________________________________________________________________________________________

```{r data preprocessing}
train_raw <- train_raw %>%
  select(
    -emp_title,
    -title,
    -address,
    -last_credit_pull_d
  )
```


### Applying Preprocessing to Holdout Data
_____________________________________________________________________________________________________________

#### Character-to-Factor Conversion

##### Term

```{r data preprocessing}
hold_raw <- hold_raw %>%
  mutate(
    term = factor(term, leveels = c("36 months", "60 months"))
  )
```

##### Employee Length

```{r data preprocessing}
hold_raw <- hold_raw %>%
  mutate(
    emp_length = forcats::fct_na_value_to_level(factor(emp_length), level = "Missing"),
    emp_length_numeric = case_when(
        emp_length == "< 1 year" ~ 0,
        emp_length == "1 year" ~ 1,
        emp_length == "2 years" ~ 2,
        emp_length == "3 years" ~ 3,
        emp_length == "4 years" ~ 4,
        emp_length == "5 years" ~ 5,
        emp_length == "6 years" ~ 6,
        emp_length == "7 years" ~ 7,
        emp_length == "8 years" ~ 8,
        emp_length == "9 years" ~ 9,
        emp_length == "10+ years" ~ 10,
        TRUE ~ NA_real_
      )
  )
```

##### Home Ownership

```{r data preprocessing}
hold_raw$home_ownership <- as.factor(hold_raw$home_ownership)
```

##### Verification Status

```{r data preprocessing}
hold_raw$verification_status <- as.factor(hold_raw$verification_status)
```

##### Purpose

```{r data preprocessing}
hold_raw$purpose <- as.factor(hold_raw$purpose)
```

##### Initial List Status

```{r data preprocessing}
hold_raw$initial_list_status <- as.factor(hold_raw$initial_list_status)
```

##### Application Type

```{r data preprocessing}
hold_raw$application_type <- as.factor(hold_raw$application_type)
```

#### Character-to-Numeric Conversion

##### Term Number

```{r data hold_raw}
hold_raw <- train_raw %>%
  mutate(
    term_num = readr::parse_number(as.character(term))
  )
```

##### Employee Length Number

```{r data preprocessing}
hold_raw <- hold_raw %>%
  mutate(
    emp_length_numeric = readr::parse_number(as.character(emp_length))
  )
```


#### Character-to-Date Conversion

##### Issue Date

```{r data preprocessing}
hold_raw <- trahold_rawin_raw %>%
  mutate(
    issue_d = lubridate::my(issue_d)
  )
```

##### Earliest Credit Line

```{r data preprocessing}
hold_raw <- hold_raw %>%
  mutate(
    earliest_cr_line = lubridate::my(earliest_cr_line)
  )
```

#### Missing Value Treatment

```{r data preprocessing}
num_cols_hold <- names(hold_raw)[sapply(hold_raw, is.numeric)]

train_medians <- sapply(train_raw[num_cols_train], median, na.rm = TRUE)

for (v in num_cols_hold) {
  hold_raw[[v]][!is.finite(hold_raw[[v]])] <- NA_real_
  hold_raw[[v]][is.na(hold_raw[[v]])] <- train_medians[[v]]
```

#### Handling Rare or Sparse Categories

##### Application Type

```{r data preprocessing}
hold_raw <- hold_raw %>%
  mutate(
    application_type = forcats::fct_collapse(
      factor(application_type),
      OTHER = c("JOINT", "DIRECT_PAY")
    )
  )
```

##### Home Ownership

```{r data preprocessing}
hold_raw <- hold_raw %>%
  mutate(
    home_ownership = forcats::fct_collapse(
      factor(home_ownership),
      OTHER = c("OTHER", "ANY", "NONE")
    )
  )
```

#### Removing Unnecessary Variables

```{r data preprocessing}
hold_raw <- hold_raw %>%
    select(
      -emp_title,
      -title,
      -address,
      -last_credit_pull_d
    )
```

## *Feature Engineering*

### Ordinal Risk Features
_____________________________________________________________________________________________________________

The variables `grade_score` and `sub_grade_score` translate lender risk grades into ordered numeric scores. This is useful because grade A is lower risk than grade B, grade B is lower risk than grade C, and so on.

#### Training Dataset

```{r feature engineering}
train_raw <- train_raw %>%
  mutate(
    grade_score = match(stringr::str_to_upper(stringr::str_trim(grade)), LETTERS[1:7]),
    sub_grade_score = as.integer(factor(
      sub_grade,
      levels = as.vector(outer(LETTERS[1:7], 1:5, paste0))
    ))
  )
```

#### Holdout Dataset

```{r feature engineering}
hold_raw <- hold_raw %>%
  mutate(
    grade_score = match(stringr::str_to_upper(stringr::str_trim(grade)), LETTERS[1:7]),
    sub_grade_score = as.integer(factor(
      sub_grade,
      levels = as.vector(outer(LETTERS[1:7], 1:5, paste0))
    ))
  )
```

### Sparse Count Binning
_____________________________________________________________________________________________________________

Sparse count variables such as public records, bankruptcies, recent inquiries, and delinquencies are grouped into categories. The bins `0`, `1`, and `2+` are easier to interpret and more stable than rare exact counts.

#### Training Dataset

```{r feature engineering}
train_raw <- train_raw %>%
  mutate(
    no_delinq_flag = ifelse(is.na(mths_since_last_delinq), 1, 0),
    mths_since_last_delinq = ifelse(is.na(mths_since_last_delinq), 999, mths_since_last_delinq),
    pub_rec_group = factor(case_when(
      pub_rec == 0 ~ "0",
      pub_rec == 1 ~ "1",
      pub_rec >= 2 ~ "2+",
      TRUE ~ "Missing"
    )),
    pub_rec_bankruptcies_group = factor(case_when(
      pub_rec_bankruptcies == 0 ~ "0",
      pub_rec_bankruptcies == 1 ~ "1",
      pub_rec_bankruptcies >= 2 ~ "2+",
      TRUE ~ "Missing"
    )),
    inq_last_6mths_group = factor(case_when(
      inq_last_6mths == 0 ~ "0",
      inq_last_6mths == 1 ~ "1",
      inq_last_6mths >= 2 ~ "2+",
      TRUE ~ "Missing"
    )),
    acc_now_delinq_group = factor(ifelse(acc_now_delinq == 0, "No", "Yes")),
    delinq_2yrs_group = factor(case_when(
      delinq_2yrs == 0 ~ "0",
      delinq_2yrs == 1 ~ "1",
      delinq_2yrs >= 2 ~ "2+",
      TRUE ~ "Missing"
    ))
  )
```

#### Holdout Dataset

```{r feature engineering}
hold_raw <- hold_raw %>%
  mutate(
    no_delinq_flag = ifelse(is.na(mths_since_last_delinq), 1, 0),
    mths_since_last_delinq = ifelse(is.na(mths_since_last_delinq), 999, mths_since_last_delinq),
    pub_rec_group = factor(case_when(
      pub_rec == 0 ~ "0",
      pub_rec == 1 ~ "1",
      pub_rec >= 2 ~ "2+",
      TRUE ~ "Missing"
    )),
    pub_rec_bankruptcies_group = factor(case_when(
      pub_rec_bankruptcies == 0 ~ "0",
      pub_rec_bankruptcies == 1 ~ "1",
      pub_rec_bankruptcies >= 2 ~ "2+",
      TRUE ~ "Missing"
    )),
    inq_last_6mths_group = factor(case_when(
      inq_last_6mths == 0 ~ "0",
      inq_last_6mths == 1 ~ "1",
      inq_last_6mths >= 2 ~ "2+",
      TRUE ~ "Missing"
    )),
    acc_now_delinq_group = factor(ifelse(acc_now_delinq == 0, "No", "Yes")),
    delinq_2yrs_group = factor(case_when(
      delinq_2yrs == 0 ~ "0",
      delinq_2yrs == 1 ~ "1",
      delinq_2yrs >= 2 ~ "2+",
      TRUE ~ "Missing"
    ))
  )
```

### Credit History Features
_____________________________________________________________________________________________________________

The variable `fico_score` combines the lower and upper FICO range bounds into an approximate average score. `credit_history_years` measures the length of credit history between the earliest credit line and loan issue date.

#### Training Dataset

```{r feature engineering}
train_raw <- train_raw %>%
  mutate(
    fico_score = (fico_range_low + fico_range_high) / 2,
    credit_history_days = as.numeric(issue_d - earliest_cr_line),
    credit_history_years = credit_history_days / 365.25
  )
```

#### Holdout Dataset

```{r feature engineering}
hold_raw <- hold_raw %>%
  mutate(
    fico_score = (fico_range_low + fico_range_high) / 2,
    credit_history_days = as.numeric(issue_d - earliest_cr_line),
    credit_history_years = credit_history_days / 365.25
  )
```

### Ratio Features
_____________________________________________________________________________________________________________

#### Training Dataset

```{r feature engineering}
train_raw <- train_raw %>%
  mutate(
    installment_to_loanamt = installment / pmax(loan_amnt, 1),
    installment_to_income = installment / pmax(annual_inc, 1),
    loan_to_income = loan_amnt / pmax(annual_inc, 1)
  )
```

#### Holdout Dataset

```{r feature engineering}
hold_raw <- hold_raw %>%
  mutate(
    installment_to_loanamt = installment / pmax(loan_amnt, 1),
    installment_to_income = installment / pmax(annual_inc, 1),
    loan_to_income = loan_amnt / pmax(annual_inc, 1)
  )
```

### Final Cleanup
_____________________________________________________________________________________________________________

The cleaned dataset has no missing values after preprocessing and contains a mixture of numeric, categorical, and engineered predictors.

#### Training Dataset

```{r feature engineering}
train_raw <- train_raw %>%
  select(
    -issue_d,
    -earliest_cr_line,
    -grade,
    -sub_grade,
    -fico_range_low,
    -fico_range_high,
    -installment,
    -annual_inc
  )
```

#### Holdout Dataset

```{r feature engineering}
hold_raw <- hold_raw %>%
  select(
    -issue_d,
    -earliest_cr_line,
    -grade,
    -sub_grade,
    -fico_range_low,
    -fico_range_high,
    -installment,
    -annual_inc
  )
```

## *Re-Assignment*
____________________________________________________________________________________________________________

Since the datasets are "cleaned", the `raw` denotion is removed.

```{r assign}
# TRAINING DATASET RE-ASSIGNMENT
train <- train_raw

# HOLDOUT DATASET RRE-ASSIGNMENT
hold <- hold_raw
```

## *Dimension Reduction*

Dimension reduction is considered when a model uses many correlated or dummy-encoded predictors. PCA can reduce many numeric columns into a smaller set of components, but it also reduces interpretability. PCA is most useful for models that are sensitive to high-dimensional numeric input, such as neural networks or regularized models using a wide dummy matrix. It is less necessary for tree-based models because tree models can handle nonlinear splits and mixed predictor importance directly. PCA requires numeric predictors to be centered and scaled. Without scaling, variables measured on large numeric scales can dominate the components.

The code below corresponds to the appropriate PCA process. This code is not evaluated and, therefore, is optional. It is included for completeion.

```{r dimension reduction, eval=FALSE}
train_dummy <- fastDummies::dummy_cols(
  train,
  remove_first_dummy = TRUE,
  remove_selected_columns = TRUE
)
```

```{r dimension reduction, eval=FALSE}
hold_dummy <- fastDummies::dummy_cols(
  hold,
  remove_first_dummy = TRUE,
  remove_selected_columns = TRUE
)
```

```{r dimension reduction, eval=FALSE}
missing_hold_cols <- setdiff(names(train_dummy), names(hold_dummy))
for (v in missing_hold_cols) hold_dummy[[v]] <- 0
```

```{r dimension reduction, eval=FALSE}
extra_hold_cols <- setdiff(names(hold_dummy), names(train_dummy))
hold_dummy <- hold_dummy %>% select(-all_of(extra_hold_cols))
hold_dummy <- hold_dummy %>% select(all_of(names(train_dummy)))
```

```{r dimension reduction, eval=FALSE}`
pca_recipe <- recipes::recipe(loan_default ~ ., data = train_dummy) %>%
  recipes::step_zv(all_predictors()) %>%
  recipes::step_normalize(all_numeric_predictors()) %>%
  recipes::step_pca(all_numeric_predictors(), threshold = 0.90)
```

```{r dimension reduction, eval=FALSE}
pca_prep <- recipes::prep(pca_recipe, training = train_dummy)
```
```{r dimension reduction, eval=FALSE}
train_pca <- recipes::bake(pca_prep, new_data = train_dummy)
```

```{r dimension reduction, eval=FALSE}
hold_pca <- recipes::bake(pca_prep, new_data = hold_dummy)
```

## *Sample*

The significance of the class imbalance can be exemplified by the following explanation. If a model predicts most loans as non-default, it can obtain high accuracy while failing to identify default cases. Because defaults are the minority class, sampling and threshold tuning are considered as part of the modeling workflow.

### Class Distribution in Training Dataset 
_____________________________________________________________________________________________________________

The target variable distribution observed shows an imbalanced, which also matches previous observations found from the target variable distributions. 

```{r sample}

```

### Class Distribution in Holdout Dataset
_____________________________________________________________________________________________________________

For consistency purposes, the target variable distribution of the holdout dataset was also examined. A similar imbalance was observed as found in the training dataset target variable distriibution.

```{r sample}

```

### Sampling Strategy
_____________________________________________________________________________________________________________

The main sampling option considered is undersampling the majority class. Undersampling creates a balanced training dataset by keeping all default cases and sampling an equal number of non-default cases. This can increase recall for the default class, but it may discard useful information from the majority class. 

The model search compares full-distribution training with undersampled training. The current highest-F1 model was trained on the full training distribution, so the final selected model does not require undersampling.

```{r sample}
set.seed(123)

maj <- train %>% 
  filter(loan_default == "No")
min <- train %>% 
  filter(loan_default == "Yes")

maj_under <- maj %>% 
  sample_n(nrow(min))

train_under <- bind_rows(min, maj_under) %>%
  slice_sample(prop = 1)

prop.table(table(train_under$loan_default))
```

The holdout dataset is not undersampled. It remains in its natural distribution so final evaluation reflects the type of class balance expected on new loan records.

```{r sample}
prop.table(table(hold_clean$loan_default))
```

## *Exporting Cleaned Data For Modeling*

### Cleaned Training Dataset
_____________________________________________________________________________________________________________

```{r clean export}
# IMBALANCED TRAINING SET
saveRDS(train, file = 'train.rds')

# UNDERSAMPLED TRAINING SET
saveRDS(train_under, file = 'train_under.rds')
```

### Cleaned Holdout Dataset
_____________________________________________________________________________________________________________

```{r clean export}
saveRDS(hold, file = 'holdout.rds')
```

## *Modeling*

### Model Type: Regularized Logistic Regression
_____________________________________________________________________________________________________________

#### Model Package 

The logistic regression model used was derived from the 

#### Parameters



#### Workflow


#### Model Variants

##### L1-Regularization Logistic Regression Model


##### L2-Regularization Logistic Regression Model


##### Elastic Net Regularization Logistic Regression Model


#### Model Imports

```{r modeling}

```

### Model Type: Decision Tree Classifier
_____________________________________________________________________________________________________________

#### Model Package 


#### Parameters


#### Workflow


#### Model Variants


#### Model Imports

```{r modeling}

```

### Model Type: Random Forest Ensemble
_____________________________________________________________________________________________________________

#### Model Package 


#### Parameters


#### Workflow


#### Model Variants


#### Model Imports

```{r modeling}

```

### Model Type: XGBoost Tree Ensemble
_____________________________________________________________________________________________________________

#### Model Package 


#### Parameters


#### Workflow


#### Model Variants


#### Model Imports

```{r modeling}

```

### Model Type: LogBoost Tree Ensemble
_____________________________________________________________________________________________________________

#### Model Package 


#### Parameters


#### Workflow


#### Model Variants


#### Model Imports

```{r modeling}

```

### Model Type: Neural Network
_____________________________________________________________________________________________________________

#### Model Package 


#### Parameters


#### Workflow


#### Model Variants


#### Model Imports

```{r modeling}

```

### Evaluation
_____________________________________________________________________________________________________________

---

# **Results**


## *Model Comparison*

**Table 1.** A table showing the best model-type for each different model variants considered and corresponding evaluation metrics, such as F1 score, accuracy, precision, balanced accuracy, sensitivity/recall, and AUC.

| **Model Type** | **Model Parameters** | **Sample** | **F1 Score** | **AUC** | **Accuracy** | **Balanced Accuracy** | **Precision** | **Sensitivity/Recall** |
|:---------------|:---------------------|:-----------|:-------------|:--------|:-------------|:---------------------|:---------------|:-----------------------|
| dtc (gini)     |      | None |       |       |      |       |       |       |
| dtc (entropy)  |      |  None     |       |       |      |       |       |       |
| prunned dtc (gini) |      |  None     |       |       |      |       |       |       |
| prunned dtc (entropy) |      |   None    |       |       |      |       |       |       |
| cv dtc (gini) |      |    None   |       |       |      |       |       |       |
| cv dtc (entropy) |      |   None    |       |       |      |       |       |       |
| rf  |      |  None     |       |       |      |       |       |       |
| rf (cv) |      |    None   |       |       |      |       |       |       |
| logistic regression (L1)|      |  None     |       |       |      |       |       |       |
| logistic regression (L2)|      |   None    |       |       |      |       |       |       |
| logistic regression (Elastic Net) |       |   None   |       |       |       |      |       |       |       |
| Neural Network (rf variable selection) |        | None     |       |       |       |      |       |       |       |
| Neural Network (logistic regression variable selection) |      | None     |       |       |       |      |       |       |       |
| dtc (gini)     |      | Balanced |       |       |      |       |       |       |
| dtc (entropy)  |      |  Balanced     |       |       |      |       |       |       |
| prunned dtc (gini) |      |  Balanced     |       |       |      |       |       |       |
| prunned dtc (entropy) |      |   Balanced    |       |       |      |       |       |       |
| cv dtc (gini) |      |    Balanced   |       |       |      |       |       |       |
| cv dtc (entropy) |      |   Balanced    |       |       |      |       |       |       |
| rf  |      |  Balanced     |       |       |      |       |       |       |
| rf (cv) |      |    Balanced   |       |       |      |       |       |       |
| logistic regression (L1)|      |  Balanced     |       |       |      |       |       |       |
| logistic regression (L2)|      |   Balanced    |       |       |      |       |       |       |
| logistic regression (Elastic Net) |       |   Balanced   |       |       |       |      |       |       |       |
| Neural Network (rf variable selection) |        | Balanced     |       |       |       |      |       |       |       |
| Neural Network (logistic regression variable selection) |      | Balanced     |       |       |       |      |       |       |       |



---

# **Discussion**

---

# **Conclusion**

---

# **References**