---
title: "Binary Classification of Loan Default Using Machine Learning"
author: "storeraa17"
data: "`r format(Sys.Date(), '%B %d, %Y')`"
abstract: |
    The abstract should be made of
    - purpose
    - main methods
    - experimental results
    - key discussion points
    - final conclusions
keywords:
    - Predictive Statistics
    - Binary Classification
    - Machine Learning
    - Model Evaluation
    - Data Cleaning
    - Data Processing
    - Data Visualization
output:
  html_document:
    toc: true
    toc_depth: 7
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: hide
    df_print: paged
    fig_width: 8
    fig_height: 5
    fig_caption: true
    self_contained: true
  word_document:
    toc: true
    fig_caption: true
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=True,
                      message=FALSE,
                      warning=FAALSE)

knitr::opts_chunk$set(fig_width=6,
                      fig_height=4,
                      #fig_caption=not_applicable_yet
                      )

if(required(pacman)==0)
    {install.packaghes("pacman)}

pacman::p_load(
    caret,
    corrplot,
    cluster,
    DataExplorer,
    devtools,
    doParallel,
    dplyr,
    dslabs,
    e1071,
    fastDummies,
    forcats,
    GGally,
    ggplot2,
    ggthemes,
    ggrepel,
    janitor,
    leaps,
    lubridate,
    magrittr,
    mlba,
    pacman,
    recipes,
    rpart,
    rpart.plot,
    ROCR,
    purrr,
    scales,
    sf,
    skimr,
    tidycensus,
    tidygeocoder,
    tidyverse,
    tigris,
    zipcodeR
    # Add more packages as needed here
)

set.seed(123)

project_dir <- "ML_portfolio/R/loan_default/data"
```

---

# **Introduction**

The purpose of this project is to support better loan risk assessment. A reliable default prediction model can help lenders make more informed decisions, reduce financial losses, and improve consistency in loan evaluation. Rather than relying only on manual review or a limited set of business rules, a predictive model can evaluate many borrower and loan characteristics at the same time. Thus, the baility to predict loan default is a critical problem. For example, financial analysts may need to estimate the risk that a borrow will fail to repay a loan. When a loan is issued, the lender accepts financial risk based on information about the borrower, the loan terms, credit history, income, debt obligations, and application characteristics. If risk is underestimated, the lender may approve loans that are more likely to default. If risk is overestimated, the lender may reject borrowers who could have successfully repaid their loans.  Herein, we report a machine learning model used for predictive binary classification of loan default.

The machine learning model used to evaluated will be subjected to several inputs to supply the desired loan default classification output. The make-up inputs cover several important features to consider, such as borrow information, loan terms, finances, credit history, accounting, delinquencies, administration, and status variables. The data representing these variables is split into two datasets:

1. `train.csv`
2. 'holdout.csv`

The dataset 'train.csv` represents the training data and will be the dataset used for model learning. While, `holdout.csv` will be the dataset that the model is evaluated on. The evaluation of the `holdout.csv` dataset will be done on the premise of F1 score primarily. However, F1 score is not exclusively the only metric used to determine model performance. Secondary evaluation metrics used to examine model performance include accuracy, balanced accuracy, recall/sensitivity, precision, number of false positives, number of false negatives, and AUC of an ROC curve.

These performance metrics will be used to evaluate decision trees, tree ensembles, logistic regression, and neural network models. The constructions of these models can be for in the `model` folder. Nevertheless, the best resulting models, based on F1 score, will be imported and model performances will be compared to determine the optimal model.

The variables found in `train.csv` and `holdout.csv` are described in the data dictionary below:

| Variable | Type | Description |
|:---|:---|:---|
| `loan_default` | Categorical | Indicates whether the borrower defaulted. Values are `Yes` and `No`. |
| `emp_title` | Categorical / text | Borrower's reported job title. High-cardinality text variable. |
| `emp_length` | Categorical / ordinal | Borrower's reported employment length. |
| `home_ownership` | Categorical | Borrower's home ownership status. |
| `annual_inc` | Numeric | Borrower's self-reported annual income. |
| `verification_status` | Categorical | Whether income or financial information was verified. |
| `address` | Categorical / text | Borrower's recorded address. High-cardinality text variable. |
| `loan_amnt` | Numeric | Original loan principal amount. |
| `term` | Categorical | Scheduled repayment length, generally 36 or 60 months. |
| `int_rate` | Numeric | Annual interest rate assigned to the loan. |
| `installment` | Numeric | Scheduled monthly payment. |
| `grade` | Categorical / ordinal | Broad lender-assigned credit grade. |
| `sub_grade` | Categorical / ordinal | More detailed credit grade. |
| `purpose` | Categorical | Borrower's stated loan purpose. |
| `title` | Categorical / text | Borrower-provided title for the loan request. |
| `dti` | Numeric | Debt-to-income ratio. |
| `earliest_cr_line` | Date-like character | Date of borrower's earliest reported credit line. |
| `fico_range_low` | Numeric | Lower bound of reported FICO range. |
| `fico_range_high` | Numeric | Upper bound of reported FICO range. |
| `inq_last_6mths` | Numeric | Number of credit inquiries in the last six months. |
| `mths_since_last_delinq` | Numeric | Months since most recent delinquency. Missing values may indicate no prior delinquency or unavailable information. |
| `last_credit_pull_d` | Date-like character | Most recent credit pull date. |
| `open_acc` | Numeric | Number of open credit accounts. |
| `pub_rec` | Numeric | Number of derogatory public records. |
| `revol_bal` | Numeric | Total revolving balance. |
| `revol_util` | Numeric | Revolving credit utilization rate. |
| `total_acc` | Numeric | Total number of credit accounts. |
| `mort_acc` | Numeric | Number of mortgage accounts. |
| `pub_rec_bankruptcies` | Numeric | Number of public-record bankruptcies. |
| `delinq_2yrs` | Numeric | Number of delinquency incidents in the last two years. |
| `acc_now_delinq` | Numeric | Number of accounts currently delinquent. |
| `issue_d` | Date-like character | Month and year the loan was issued. |
| `initial_list_status` | Categorical | Initial listing status assigned to the loan. |
| `application_type` | Categorical | Individual, joint, or direct-pay application type. |
| `hardship_flag` | Categorical | Indicates hardship arrangement status. |
| `debt_settlement_flag` | Categorical | Indicates debt-settlement status. |

---

# **Methods**

## *Data Loading*

The datasets, `train.csv` and `holdout.csv`, were loaded and the target variable. `loan_default`, was immediately converted into a factor.

```{r data loading}
train_raw <- readr::read_csv(
  file.path(project_dir, "train.csv"),
  show_col_types = FALSE) %>%
  mutate(loan_default = factor(loan_default, levels = c("No", "Yes")))
```

```{r data loading}
hold_raw <- readr::read_csv(
  file.path(project_dir, "holdout.csv"),
  show_col_types = FALSE) %>%
  mutate(loan_default = factor(loan_default, levels = c("No", "Yes")))
```

## *Data Inspection*

### Previewing the Dataset
____________________________________________________________________________________________________________

The structure of the datasets were assessed using the `str()` function. Subsequent previewing of the datasets was done through the utilization of the `head()` function, which shows the first few rows of a dataset.

```{r data inspection}
str(train_raw)
```

```{r data inspection}
str(hold_raw)
```

```{r data inspection}
head(train_raw)
```

```{r data inspection}
head(hold_raw)
```

### Dataset Dimensions
____________________________________________________________________________________________________________

The dimensions of the dataset were determined by employed the `dim()` function.

```{r data inspection}
dim(train_raw)
```

```{r data inspection}
dim(hold_raw)
```

### The Number of Unique Instances
____________________________________________________________________________________________________________
The number of unique instances for each variable were examined via `sapply()` and `n_distinct()` functions.

```{r data inspection}
sapply(train_raw, n_distinct)
```

```{r data inspection}
sapply(hold_raw, n_distinct)
```

### Class Comparison of Variables
____________________________________________________________________________________________________________
It is important to compare the classes of each variable in the training and holdout datasets to ensure consistency. This can be done using the `sapply()` and `class()` functions. To present the findings more cleanly, a tibble can be created to display the variable names, their corresponding classes in the training dataset, and their classes in the holdout dataset. The function `tibble::tibble()` can be used to create this tibble, and the `sapply()` function can be used to apply the `class()` function to each variable in both datasets. The resulting tibble will provide a clear comparison of the variable classes between the two datasets.

```{r data inspection}
class_comparison <- tibble::tibble(
  Variable = names(train_raw),
  Train_Class = sapply(train_raw, function(x) class(x)[1]),
  Holdout_Class = sapply(hold_raw, function(x) class(x)[1])
)

class_comparison
```

## *Data Summaries & Visualizations*

### Summary Statistics
____________________________________________________________________________________________________________

The summary statistics of the datasets were examined using the `skimr::skim()` function. This function provides a comprehensive overview of the data, including the number of missing values, mean, median, standard deviation, and other relevant statistics for each variable.

```{r data processing}
skimr::skim(train_raw)
```

```{r data processing}
skimr::skim(hold_raw)
```

### Grouping Numeric Variables
____________________________________________________________________________________________________________

The numeric variables in the training dataset were identified and grouped together for further analysis. This was done using the `select()` function from the `dplyr` package, along with the `where()` function to filter for numeric variables. The names of these numeric variables were then extracted using the `names()` function.

```{r data processing}
numeric_vars <- train_raw %>%
  select(where(is.numeric)) %>%
  names()

numeric_vars
```

The numeric variables in the holdout dataset were also identified and grouped together for further analysis. This was done using the `select()` function from the `dplyr` package, along with the `where()` function to filter for numeric variables. The names of these numeric variables were then extracted using the `names()` function.

```{r data processing}
hold_numeric_vars <- hold_raw %>%
  select(where(is.numeric)) %>%
  names()

hold_numeric_vars
```

### Grouping Categorical Variables
____________________________________________________________________________________________________________

The categorical variables in the training dataset were identified and grouped together for further analysis. This was done using the `select()` function from the `dplyr` package, along with the `where()` function to filter for character or factor variables. The names of these categorical variables were then extracted using the `names()` function.

```{r data processing}
categorical_vars <- train_raw %>%
  select(where(~ is.character(.x) || is.factor(.x))) %>%
  names()

categorical_vars
```

The categorical variables in the holdout dataset were also identified and grouped together for further analysis. This was done using the `select()` function from the `dplyr` package, along with the `where()` function to filter for character or factor variables. The names of these categorical variables were then extracted using the `names()` function.

```{r data processing}
hold_categorical_vars <- hold_raw %>%
  select(where(~ is.character(.x) || is.factor(.x))) %>%
  names()

hold_categorical_vars
```

### Target Variable Distribution
____________________________________________________________________________________________________________

#### Distribution of the Target Variable in the Training Dataset

The distribution of the target variable, `loan_default`, in the training dataset was examined using the `table()` function to count the occurrences of each class, and the `prop.table()` function to calculate the proportions. The results were then visualized using a bar plot created with the `ggplot2` package.

```{r data processing}
train_target_distribution <- table(train_raw$loan_default)
train_target_proportions <- prop.table(train_target_distribution)
```

```{r data processing}
train_target_distribution_df <- as.data.frame(train_target_distribution)
colnames(train_target_distribution_df) <- c("loan_default", "Count")
```

```{r data processing}
ggplot(train_target_distribution_df, aes(x = loan_default, y = Count, fill = loan_default)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = c("No" = "#1f77b4", "Yes" = "#ff7f0e")) +
    labs(title = "Distribution of Loan Default in Training Dataset",
        x = "Loan Default",
        y = "Count") +
    theme_bw(base_size = 14) +
    theme(legend.position = "none",
          plot.title = element_text(face = "bold", hjust = 0.5))
```

#### Distribution of the Target Variable in the Holdout Dataset

The distribution of the target variable, `loan_default`, in the holdout dataset was examined using the `table()` function to count the occurrences of each class, and the `prop.table()` function to calculate the proportions. The results were then visualized using a bar plot created with the `ggplot2` package.

```{r data processing}
hold_target_distribution <- table(hold_raw$loan_default)
hold_target_proportions <- prop.table(hold_target_distribution)
```

```{r data processing}
hold_target_distribution_df <- as.data.frame(hold_target_distribution)
colnames(hold_target_distribution_df) <- c("loan_default", "Count")
```

```{r data processing}
ggplot(hold_target_distribution_df, aes(x = loan_default, y = Count, fill = loan_default)) +
    geom_bar(stat = "identity", color="white", alpha=0.9) +
    scale_fill_manual(values = c("No" = "#1f77b4", "Yes" = "#ff7f0e")) +
    labs(title = "Distribution of Loan Default in Holdout Dataset",
        x = "Loan Default",
        y = "Count") +
    theme_bw(base_size = 14) +
    theme(legend.position = "none",
          plot.title = element_text(face = "bold", hjust = 0.5))    
```

### Numeric Variable Distributions
____________________________________________________________________________________________________________

The numeric variables in the training and holdout datasets were examined using histograms to visualize their distributions. The `DataExplorer` package was used to create the histograms, with customizations applied to the theme and appearance of the plots.

#### Historgrams of Numeric Variables in the Training Dataset

```{r data processing}
DataExplorer::plot_histogram(
  train_raw %>% select(all_of(numeric_vars)),
  ggtheme =
    theme_bw(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "#1F77B4"),
      strip.text = element_text(color = "white", face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey90")
    ),
  geom_histogram_args = list(
    fill = "#4E79A7",
    color = "white",
    alpha = 0.9
  ),
  ncol = 3
)
```

#### Historgrams of Numeric Variables in the Holdout Dataset

```{r data processing}
DataExplorer::plot_histogram(
  hold_raw %>% select(all_of(hold_numeric_vars)),
  ggtheme =
    theme_bw(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "#a3231e"),
      strip.text = element_text(color = "white", face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey90")
    ),
  geom_histogram_args = list(
    fill = "#a3231e",
    color = "white",
    alpha = 0.9
  ),
  ncol = 3
)
```

### Categorical Variable Distributions
____________________________________________________________________________________________________________

#### Frequecy Tables of Categorical Variables in the Training Dataset

```{r data processing}
for (col in names(categorical_vars)) {
    print(table(categorical_vars[[col]])[1:5])
}
```

#### Frequecy Tables of Categorical Variables in the Holdout Dataset

```{r data processing}
for (col in names(hold_categorical_vars)) {
    print(table(hold_categorical_vars[[col]])[1:5])
}
```


---

# **Results**

---

# **Discussion**

---

# **Conclusion**

---

# **References**