---
title: "Logistic Regression Modeling for Predicting Loan Default"
author: "storera17"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: show
    df_print: paged
    number_sections: yes
    theme: readable
    toc: yes
    toc_float: yes
    code_download: yes
  word_document:
    toc: no
params:
  response_var: "loan_default"
  id_var: ["load_id"] #identifier to be dropped

---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE)
knitr::opts_chunk$set(fig_width=6,fig_height=4)

if(require(pacman)==0)
    {install.packages("pacman")}

pacman::p_load(
    caret,
    corrplot,
    cluster,
    DataExplorer,
    devtools,
    doParallel,
    dplyr,
    dslabs,
    e1071,
    fastDummies,
    forcats,
    GGally,
    ggplot2,
    ggthemes,
    ggrepel,
    janitor,
    leaps,
    lubridate,
    magrittr,
    mlba,
    pacman,
    recipes,
    rpart,
    rpart.plot,
    ROCR,
    purrr,
    scales,
    sf,
    skimr,
    tidycensus,
    tidygeocoder,
    tidyverse,
    tigris,
    zipcodeR
    # Add more packages as needed here
)

set.seed(123)

project_dir <- "ML_portfolio/R/loan_default/data"
```

# **Read in the Data**

```{r load}
train <- readRDS("train.rds")
train_under <- readRDS("train_under.rds")
hold <- readRDS("holdout.rds")
```

# **Skim the Data**

```{r skim}
skimr::skim(train)
```

```{r skim}
skimr::skim(train_under)
```

```{r skim}
skimr::skim(hold)
```

# **Modeling**

## *Explanation of the Parameters*

### `train()` Parameters
____________________________________________________________________________________________________________

The `carat` package contains the function `train()` that sets up a grid of tuning parameters for a number of classification and regression routines, fits each model, and calculates a resampling based performance measure. This function fits predictive models over different tuning parameters. The different tuning parameters are described in **Table 1**. These parameters were utilized to perform logistic regression. The particular parameter that enables logistic regression in the `train()` function is `method`. Setting `method` to `glmnet` runs a logistic regression model.

**Table 1.** A tabulation of the `train()` function's parameters and corresponding descriptions of the parameters.

| Parameter | Description |
|:---|:---|
| `x` | Predictor data. For the default method, rows are samples and columns are features. This can be a matrix, data frame, sparse matrix, or similar object, but columns must be named. If using the recipe method, `x` should be an unprepared `recipes::recipe()` object that defines the outcome, predictors, and preprocessing steps. |
| `...` | Additional arguments passed to the underlying classification or regression routine. Tuning parameters should not be passed here because they belong in `tuneGrid` or `tuneLength`. |
| `y` | Outcome vector for each sample. This should be numeric for regression or a factor for classification. |
| `method` | String specifying the model type to train, such as `"glmnet"`, `"rpart"`, `"ranger"`, or `"nnet"`. Available model names can be checked with `names(getModelInfo())`. |
| `preProcess` | Character vector specifying preprocessing steps for predictor data, such as `"center"`, `"scale"`, `"range"`, `"medianImpute"`, `"pca"`, or `"YeoJohnson"`. This applies only when `x` is a matrix or data frame, not when using a recipe. |
| `weights` | Numeric vector of case weights. This only affects models that support weighted observations. |
| `metric` | Summary metric used to select the optimal model. Common defaults are `"RMSE"` and `"Rsquared"` for regression or `"Accuracy"` and `"Kappa"` for classification. If using a custom `summaryFunction`, `metric` should match one of the returned metric names. |
| `maximize` | Logical value indicating whether the selected metric should be maximized or minimized. |
| `trControl` | A `trainControl()` object that defines resampling, class probabilities, saved predictions, summary functions, sampling, seeds, and parallel behavior. |
| `tuneGrid` | Data frame of tuning parameter combinations. Column names must match the model’s tuning parameters, which can be checked with `modelLookup()` or `getModelInfo()`. |
| `tuneLength` | Integer controlling the size or granularity of the automatically generated tuning grid. If random search is used, this is the maximum number of tuning combinations. |
| `form` | Formula specifying the model structure, usually in the form `outcome ~ predictor1 + predictor2 + ...`. |
| `data` | Data frame containing the variables referenced in the formula or recipe. |
| `subset` | Index vector specifying which rows should be used in the training sample. |
| `na.action` | Function specifying what to do when missing values are found. The default is to fail; alternatives such as `na.omit` remove incomplete cases. |
| `contrasts` | List specifying contrasts for factor variables used in the model formula. |

Not all of the parameters described in **Table 1** will be utilized.

### `trainControl()` Parameters
____________________________________________________________________________________________________________

An explanation of the parameters in the `trainControl` object are described in **Table 2**.

**Table 2.** A tabulation of the parameters in `trainControl()` and their corresponding descriptions.

| Parameter | Description |
|:---|:---|
| `method` | The resampling method, such as `"boot"`, `"cv"`, `"repeatedcv"`, `"LOOCV"`, `"LGOCV"`, `"none"`, `"oob"`, `"timeslice"`, `"adaptive_cv"`, `"adaptive_boot"`, or `"adaptive_LGOCV"`. |
| `number` | The number of folds or resampling iterations. |
| `repeats` | For repeated k-fold cross-validation, the number of complete fold sets to compute. |
| `p` | For leave-group-out cross-validation, the training percentage. |
| `search` | Either `"grid"` or `"random"`, describing how the tuning parameter grid is determined. |
| `initialWindow`, `horizon`, `fixedWindow`, `skip` | Arguments passed to `createTimeSlices()` when `method = "timeslice"`. |
| `verboseIter` | Logical value indicating whether a training log should be printed. |
| `returnData` | Logical value indicating whether the training data should be saved. |
| `returnResamp` | Controls how much resampled summary metric output is saved. Values are `"final"`, `"all"`, or `"none"`. |
| `savePredictions` | Controls how much holdout prediction output from each resample is saved. Values are `"all"`, `"final"`, or `"none"`. The `"final"` option saves predictions for the optimal tuning parameters. |
| `classProbs` | Logical value indicating whether class probabilities should be computed for classification models. |
| `summaryFunction` | Function used to compute performance metrics across resamples. |
| `selectionFunction` | Function used to select the optimal tuning parameter. |
| `preProcOptions` | List of options passed to `preProcess()`. The preprocessing method itself is specified in the `preProc` argument of `train()`. |
| `sampling` | Optional sampling method applied after resampling, usually for class imbalance. Common values include `"none"`, `"down"`, `"up"`, `"smote"`, and `"rose"

### `tuneGrid()` Parameters
___________________________________________________________________________________________________________

The `tuneGrid` arguement utilizes the `expand.grid()` function with two nested `seq()` functions. Together, these functions work as a grid search for specific parameters. 

- alpha = 1: Sets the model to LASSO regression.
- tuneGrid: This creates a grid of lambda values for cv.
  - Small lambda → less penalty → model behaves more like plain logistic regression.
  - Large lambda → more shrinkage → more coefficients get pushed toward zero.


## *Logistic Regression Model - Lasso Regularization*

### Determine Parameters
____________________________________________________________________________________________________________

#### Set Up `trainControl()`

```{r lasso}
ctrl <- trainControl(method = "cv", 
                     number = 4,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE,
                     savePredictions = "final")
```

#### Set Up `tuneGrid()`

```{r lasso}
grid <- expand.grid(
  alpha = seq(0, 1, length = 10),
  lambda = seq(0.001, 0.1, length = 5)
)
```


### Model Fitting
____________________________________________________________________________________________________________

#### Fit Model on Imbalanced Training Dataset

```{r lasso}
log_lasso <- caret::train(
  loan_default ~ .,
  data = train,
  method = "glmnet",
  trControl = ctrl,
  tuneGrid = grid,
  metric = "ROC",
  family = "binomial")
```

#### Fit Model on Undersampled Training Dataset

```{r lasso}
log_lasso_under <- caret::train(
  loan_default ~ .,
  data = train_under,
  method = "glmnet",
  trControl = ctrl,
  tuneGrid = grid,
  metric = "ROC",
  family = "binomial")
```

### Predict on Holdout Dataset
____________________________________________________________________________________________________________



### Evaluate
____________________________________________________________________________________________________________



### Store & Export
____________________________________________________________________________________________________________



## *Logistic Regression Model - Ridge Regularization*

### Determine Optimal Parameters
____________________________________________________________________________________________________________



### Fit Model
____________________________________________________________________________________________________________



### Predict on Holdout Dataset
____________________________________________________________________________________________________________



### Evaluate
____________________________________________________________________________________________________________



### Store & Export
____________________________________________________________________________________________________________



## *Logistic Regression Model - Elastic Net Regularization*

### Determine Optimal Parameters
____________________________________________________________________________________________________________



### Fit Model
____________________________________________________________________________________________________________



### Predict on Holdout Dataset
____________________________________________________________________________________________________________



### Evaluate
____________________________________________________________________________________________________________



### Store & Export
____________________________________________________________________________________________________________



# *Logistic Regression Model - Lasso Regularization - Undersample*

### Determine Optimal Parameters
____________________________________________________________________________________________________________



### Fit Model
____________________________________________________________________________________________________________



### Predict on Holdout Dataset
____________________________________________________________________________________________________________



### Evaluate
____________________________________________________________________________________________________________



### Store & Export
____________________________________________________________________________________________________________



## *Logistic Regression Model - Ridge Regularization - Undersample*

### Determine Optimal Parameters
____________________________________________________________________________________________________________



### Fit Model
____________________________________________________________________________________________________________



### Predict on Holdout Dataset
____________________________________________________________________________________________________________



### Evaluate
____________________________________________________________________________________________________________



### Store & Export
____________________________________________________________________________________________________________



## *Logistic Regression Model - Elastic Net Regularization - Undersample*

### Determine Optimal Parameters
____________________________________________________________________________________________________________



### Fit Model
____________________________________________________________________________________________________________



### Predict on Holdout Dataset
____________________________________________________________________________________________________________



### Evaluate
____________________________________________________________________________________________________________



### Store & Export
____________________________________________________________________________________________________________



### Fit the Model

The original parameters are commented out such that we could explore the optimal values. We first used the seq() function and got values for alpha and lambda as 0.25 and 0.001, respectively. Then, we set lambda to be 0.001 and let optimized our alpha value and got it to be 0.11111111. Finally, we optimized lambda, just to double-check, and at a set 0.11111111 alpha value, the best lambda we got was 0.001.

```{r}
logit_fit <- caret::train(
  loan_default ~ .,
  data = train,
  method = "glmnet",
  trControl = ctrl,
  tuneGrid = expand.grid(
    alpha = seq(0, 1, length=8),
    lambda = seq(.001,.1, length=5)),
  metric = "ROC",
  family = "binomial"
)
```

```{r}
#assigned_variable <- caret::train(
#  target_variable ~ .,
#  data = dataset_to_be_used
#  method = "model_type_used"
#
#)
```

**Explanation of the tuneGrid argument**

- alpha = 1: Sets the model to LASSO regression.
- tuneGrid: This creates a grid of lambda values for cv.
  - Small lambda → less penalty → model behaves more like plain logistic regression.
  - Large lambda → more shrinkage → more coefficients get pushed toward zero.

**Below is a graphical exploration of the model to find the optimal lambda value**

The plot generated yields a the various ROC (Cross-Validation) values according to the regularization paramter. Thus, we can use this plot to determine our optimal regularization parameter.

```{r}
plot(logit_fit)
```

This table tells us the optimal lambda and alpha values. We previously stated these values above.

```{r}
logit_fit$bestTune
```

Here, we see a more comprehensive examination of different alpha and lambda values corresponding to performance metrics of the model they generate. In this case, lambda and alpha were selected using the F1 score because we set the metric to F1.

```{r}
logit_fit$results
```

**Below we look at the coefficients of the model**

```{r}
coef(logit_fit$finalModel,logit_fit$bestTune$lambda)
```

## Predict on the Test Set

Here, we created a dataframe that includes the predictions on the test data along with the actual values. We first create the dataframe with just the loan_default response variable. Then, we do the predictions on the holdout dataset (probabilitiy and classification). Lastly, we add predictions to the dataframe.

**Predict on Holdout**
```{r}
hold_pred <- data.frame(
  Predicted_Loan_Default = hold$loan_default,
 # pred_class = predict(logit_fit,
  #                   newdata = hold,
  #                   type = "raw",
 #                    )[,"Yes"],
  pred_prob = predict(logit_fit,
                      newdata = hold,
                      type = "prob")[,"Yes"])
hold_pred <- hold_pred %>%
  mutate(pred_class = factor(ifelse(hold_pred$pred_prob > 0.15,
                                         "Yes", "No"),
                                  levels=c("Yes","No")))
head(hold_pred)
```


## Evaluate the Model
### Confusion Matrix

```{r}
confusionMatrix(hold_pred$pred_class,
                hold_pred$Predicted_Loan_Default,
                positive = "Yes")
```

### ROC Curve
The generation of the ROC curve is done via the two chunks below. Additionally, the corresponding AUC and F1 value are reported below. This is only done on the holdout data.

**Compute the ROC curve**

```{r}
preds <- ROCR::prediction(hold_pred$pred_prob, hold_pred$Predicted_Loan_Default)
perf <- ROCR::performance(preds,"tpr","fpr")
```

**Plot Curve**
```{r}
plot(perf, col = "blue", lwd = 2, main = "ROC Curve fo rOur Model")
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)
```

**Extract AUC**
```{r}
auc <- ROCR::performance(preds, "auc")@y.values[[1]]
print(paste("The AUC is ",round(auc,2)))
```

### F1 Score
**Holdout**
The F1 score was obtained using the F1_Score() function from the MLmetrics package. The F1 score for our model was 0.42 for imbalance data.
```{r}
F1 <- MLmetrics::F1_Score(y_pred = hold_pred$pred_class,
                          y_true = hold$loan_default,
                          positive = "Yes")


print(paste("The F1 score is ",round(F1,2)))
```

# Balanced Dataset
## Undersample
Here, we undersample the data, such the the number for each class is equivalent.
```{r}
set.seed(123)
maj <- train %>% dplyr::filter(loan_default == "No")
min <- train %>% dplyr::filter(loan_default == "Yes")
maj_under <- maj %>% dplyr::sample_n(nrow(min))
train_us <- dplyr::bind_rows(min, maj_under) %>% dplyr::slice_sample(prop = 1)
prop.table(table(train_us$loan_default))
```


### Fit the Model
We  used the seq() function to the optimal values for alpha and lambda, respectively.

**Explanation of the tuneGrid argument**
- alpha = 1: Sets the model to LASSO regression.
- tuneGrid: This creates a grid of lambda values for cv.
  - Small lambda → less penalty → model behaves more like plain logistic regression.
  - Large lambda → more shrinkage → more coefficients get pushed toward zero.

**Below is a graphical exploration of the model to find the optimal lambda value**
The plot generated yields a the various ROC (Cross-Validation) values according to the regularization paramter. Thus, we can use this plot to determine our optimal regularization parameter.
```{r}
plot(logit_fit_us)
```

This table tells us the optimal lambda and alpha values. We previously stated these values above.
```{r}
logit_fit_us$bestTune
```

Here, we see a more comprehensive examination of different alpha and lambda values corresponding to performance metrics of the model they generate. In this case, lambda and alpha were selected using the F1 score because we set the metric to F1.
```{r}
logit_fit_us$results
```

**Below we look at the coefficients of the model**
```{r}
coef(logit_fit_us$finalModel,logit_fit_us$bestTune$lambda)
```

## Predict on the Test Set
Here, we created dataframes that includes the predictions on the training and test data along with the actual values. We first create the dataframe with just the loan_default response variable labeled as Predicted_Loan_Defualt. Then, we do the predictions on the training and holdout datasets (probabilitiy and classification). Lastly, we add these predictions to the corresponding dataframes.

**Predict on Holdout**
```{r}
hold_pred_us <- data.frame(
  Predicted_Loan_Default = hold$loan_default,
  pred_prob = predict(logit_fit_us,
                      newdata = hold,
                      type = "prob")[, "Yes"])
hold_pred_us <- hold_pred_us %>%
  mutate(pred_class = factor(ifelse(hold_pred_us$pred_prob > 0.45,
                                          "Yes",
                                          "No")))
```


## Evaluate the Model
### Confusion Matrix

```{r}
confusionMatrix(hold_pred_us$pred_class,
                hold_pred_us$Predicted_Loan_Default,
                positive = "Yes")
```

### ROC Curve
The generation of the ROC curve is done via the two chunks below. Additionally, the corresponding AUC and F1 value are reported below.

**Compute the ROC curve**
```{r}
preds_us <- ROCR::prediction(hold_pred_us$pred_prob,
                             hold_pred_us$Predicted_Loan_Default)
perf_us <- ROCR::performance(preds_us, "tpr", "fpr")
```

**Plot Curve**
```{r}
plot(perf_us, col = "blue", lwd = 2, main = "ROC Curve fo rOur Model")
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)
```

**Extract AUC**
Here, we get the auc value.
```{r}
auc_us <- ROCR::performance(preds_us, "auc")@y.values[[1]]
print(paste("The AUC is ",round(auc_us,2)))
```

### F1 Score
```{r}
F1_us <- MLmetrics::F1_Score(y_pred = hold_pred_us$pred_class,
                            y_true = hold$loan_default,
                            positive = "Yes")


print(paste("The F1 score is ",round(F1_us,2)))
```

# Conclusion
The undersampled and original datasets showed similar F1 scores. However, the sensitivity for the model trained using the original dataset. Thus, the better dataset would be considered to be that trained with the whole dataset because sensitivity reflects the ability of the model to capture the positive class, which in this case is significantly more influential than capturing the negative class.

**New Imbalance Dataset**
| threshold | F1 Score |
|-----------|----------|
|  0.8     |  0.13    |
|  0.7     |  0.13    |
| 0.5      |  0.24        |
| 0.3      |  0.43        |
| 0.2      |  0.45        |
| 0.15      |    0.43      |

**New Imbalance Dataset**
| threshold | F1 Score |
|-----------|----------|
|  0.5     |  0.45    |
|  0.4      |  0.43    |
| 0.3      |  0.        |
| 0.6      |  0.45        |
| 0.7      |  0.44        |
| 0.8      |    0.43      |
