---
title: "Logistic Regression Modeling for Predicting Loan Default"
author: "storera17"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: show
    df_print: paged
    number_sections: yes
    theme: readable
    toc: yes
    toc_float: yes
    code_download: yes
  word_document:
    toc: no
params:
  response_var: "loan_default"
  id_var: ["load_id"] #identifier to be dropped 

---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE)
knitr::opts_chunk$set(fig_width=6,fig_height=4)

if(require(pacman)==0)
    {install.packages("pacman")}

pacman::p_load(
    caret,
    corrplot,
    cluster,
    DataExplorer,
    devtools,
    doParallel,
    dplyr,
    dslabs,
    e1071,
    fastDummies,
    forcats,
    GGally,
    ggplot2,
    ggthemes,
    ggrepel,
    janitor,
    leaps,
    lubridate,
    magrittr,
    mlba,
    pacman,
    recipes,
    rpart,
    rpart.plot,
    ROCR,
    purrr,
    scales,
    sf,
    skimr,
    tidycensus,
    tidygeocoder,
    tidyverse,
    tigris,
    zipcodeR
    # Add more packages as needed here
)

set.seed(123)

project_dir <- "ML_portfolio/R/loan_default/data"
```

# **Read in the Data**

```{r load}
train <- readRDS("train.rds")
train_under <- readRDS("train_under.rds")
hold <- readRDS("holdout.rds")
```

# **Skim the Data**

```{r skim}
skimr::skim(train)
```

```{r skim}
skimr::skim(train_under)
```

```{r skim}
skimr::skim(hold)
```

# **Modeling**

## *Set Up trainControl Object*
____________________________________________________________________________________________________________

An explanation of the parameters in the `trainControl` object as shown below:

- method = “cv”, number = 10: We are using 10-fold cross validation. We will use this to select our hyperparameters, 𝜆 and α. 
- summaryFunction = twoClassSummary: This tells caret to use a custom summary function to evaluate model performance. twoClassSummary computes:
  - ROC AUC (Area Under the Curve),
  - Sensitivity (Recall/TPR),
  - Specificity (TNR). Note: Your target variable must be a factor with two levels, and the positive class must be the first level of the factor when using twoClassSummary in caret. This is because twoClassSummary treats the first factor level as the positive class. Always check to make sure the first level of the target is the “success” class. You can change the order of the factor levels by reordering the labels.
- classProbs = TRUE: This tells caret to compute predicted probabilities (not just class labels). Required for metrics like AUC, which need probabilities.
- savePredictions = TRUE: This tells caret to save the model’s predictions on the training data.

```{r}
ctrl <- trainControl(method = "cv", number = 4, 
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE, 
                     savePredictions = "final")
```

### Logistic Regression Model - Lasso Regularization
____________________________________________________________________________________________________________




### Logistic Regression Model - Ridge Regularization
____________________________________________________________________________________________________________



### Logistic Regression Model - Elastic Net Regularization
____________________________________________________________________________________________________________



bject as shown below:


### Fit the Model

The original parameters are commented out such that we could explore the optimal values. We first used the seq() function and got values for alpha and lambda as 0.25 and 0.001, respectively. Then, we set lambda to be 0.001 and let optimized our alpha value and got it to be 0.11111111. Finally, we optimized lambda, just to double-check, and at a set 0.11111111 alpha value, the best lambda we got was 0.001.

```{r}
logit_fit <- caret::train(
  loan_default ~ ., 
  data = train,
  method = "glmnet",
  trControl = ctrl,
  tuneGrid = expand.grid(
    alpha = seq(0, 1, length=8),
    lambda = seq(.001,.1, length=5)),
  metric = "ROC",
  family = "binomial"
)
```

**Explanation of the tuneGrid argument**

- alpha = 1: Sets the model to LASSO regression.
- tuneGrid: This creates a grid of lambda values for cv.
  - Small lambda → less penalty → model behaves more like plain logistic regression.
  - Large lambda → more shrinkage → more coefficients get pushed toward zero.

**Below is a graphical exploration of the model to find the optimal lambda value**

The plot generated yields a the various ROC (Cross-Validation) values according to the regularization paramter. Thus, we can use this plot to determine our optimal regularization parameter. 

```{r}
plot(logit_fit)
```

This table tells us the optimal lambda and alpha values. We previously stated these values above.

```{r}
logit_fit$bestTune
```

Here, we see a more comprehensive examination of different alpha and lambda values corresponding to performance metrics of the model they generate. In this case, lambda and alpha were selected using the F1 score because we set the metric to F1. 

```{r}
logit_fit$results
```

**Below we look at the coefficients of the model**

```{r}
coef(logit_fit$finalModel,logit_fit$bestTune$lambda)
```

## Predict on the Test Set

Here, we created a dataframe that includes the predictions on the test data along with the actual values. We first create the dataframe with just the loan_default response variable. Then, we do the predictions on the holdout dataset (probabilitiy and classification). Lastly, we add predictions to the dataframe. 

**Predict on Holdout**
```{r}
hold_pred <- data.frame(
  Predicted_Loan_Default = hold$loan_default,
 # pred_class = predict(logit_fit, 
  #                   newdata = hold,
  #                   type = "raw",
 #                    )[,"Yes"],
  pred_prob = predict(logit_fit, 
                      newdata = hold,
                      type = "prob")[,"Yes"])
hold_pred <- hold_pred %>% 
  mutate(pred_class = factor(ifelse(hold_pred$pred_prob > 0.15, 
                                         "Yes", "No"),
                                  levels=c("Yes","No")))
head(hold_pred)
```


## Evaluate the Model
### Confusion Matrix

```{r}
confusionMatrix(hold_pred$pred_class, 
                hold_pred$Predicted_Loan_Default, 
                positive = "Yes")
```

### ROC Curve
The generation of the ROC curve is done via the two chunks below. Additionally, the corresponding AUC and F1 value are reported below. This is only done on the holdout data. 

**Compute the ROC curve**

```{r}
preds <- ROCR::prediction(hold_pred$pred_prob, hold_pred$Predicted_Loan_Default)
perf <- ROCR::performance(preds,"tpr","fpr")
```

**Plot Curve**
```{r}
plot(perf, col = "blue", lwd = 2, main = "ROC Curve fo rOur Model")
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)
```
  
**Extract AUC**
```{r}
auc <- ROCR::performance(preds, "auc")@y.values[[1]]
print(paste("The AUC is ",round(auc,2)))
```

### F1 Score
**Holdout**
The F1 score was obtained using the F1_Score() function from the MLmetrics package. The F1 score for our model was 0.42 for imbalance data.
```{r}
F1 <- MLmetrics::F1_Score(y_pred = hold_pred$pred_class,
                          y_true = hold$loan_default,
                          positive = "Yes")


print(paste("The F1 score is ",round(F1,2)))
```

# Balanced Dataset
## Undersample
Here, we undersample the data, such the the number for each class is equivalent.
```{r}
set.seed(123)
maj <- train %>% dplyr::filter(loan_default == "No")
min <- train %>% dplyr::filter(loan_default == "Yes")
maj_under <- maj %>% dplyr::sample_n(nrow(min))
train_us <- dplyr::bind_rows(min, maj_under) %>% dplyr::slice_sample(prop = 1)
prop.table(table(train_us$loan_default))
```

## Fit a Logistic Regression Model Using Lasso Selection with caret
### Set up a trainControl object
```{r}
ctrl <- trainControl(method = "cv", number = 4, 
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE, 
                     savePredictions = "final")
```
- method = “cv”, number = 10: We are using 10-fold cross validation. We will use this to select our hyperparameters, 𝜆 and α. 
- summaryFunction = twoClassSummary: This tells caret to use a custom summary function to evaluate model performance. twoClassSummary computes:
  - ROC AUC (Area Under the Curve),
  - Sensitivity (Recall/TPR),
  - Specificity (TNR). Note: Your target variable must be a factor with two levels, and the positive class must be the first level of the factor when using twoClassSummary in caret. This is because twoClassSummary treats the first factor level as the positive class. Always check to make sure the first level of the target is the “success” class. You can change the order of the factor levels by reordering the labels.
- classProbs = TRUE: This tells caret to compute predicted probabilities (not just class labels). Required for metrics like AUC, which need probabilities.
- savePredictions = TRUE: This tells caret to save the model’s predictions on the training data.

### Fit the Model
We  used the seq() function to the optimal values for alpha and lambda, respectively. 
```{r}
logit_fit_us <- caret::train(
  loan_default ~ ., 
  data = train_us,
  method = "glmnet",
  trControl = ctrl,
  tuneGrid = expand.grid(
    alpha = seq(0, 1, length=10), 
    lambda = seq(.001,.1, length=5)), 
  metric = "ROC",
  family = "binomial"
)
```
**Explanation of the tuneGrid argument**
- alpha = 1: Sets the model to LASSO regression.
- tuneGrid: This creates a grid of lambda values for cv.
  - Small lambda → less penalty → model behaves more like plain logistic regression.
  - Large lambda → more shrinkage → more coefficients get pushed toward zero.

**Below is a graphical exploration of the model to find the optimal lambda value**
The plot generated yields a the various ROC (Cross-Validation) values according to the regularization paramter. Thus, we can use this plot to determine our optimal regularization parameter. 
```{r}
plot(logit_fit_us)
```

This table tells us the optimal lambda and alpha values. We previously stated these values above.
```{r}
logit_fit_us$bestTune
```

Here, we see a more comprehensive examination of different alpha and lambda values corresponding to performance metrics of the model they generate. In this case, lambda and alpha were selected using the F1 score because we set the metric to F1. 
```{r}
logit_fit_us$results
```

**Below we look at the coefficients of the model**
```{r}
coef(logit_fit_us$finalModel,logit_fit_us$bestTune$lambda)
```

## Predict on the Test Set
Here, we created dataframes that includes the predictions on the training and test data along with the actual values. We first create the dataframe with just the loan_default response variable labeled as Predicted_Loan_Defualt. Then, we do the predictions on the training and holdout datasets (probabilitiy and classification). Lastly, we add these predictions to the corresponding dataframes. 

**Predict on Holdout**
```{r}
hold_pred_us <- data.frame(
  Predicted_Loan_Default = hold$loan_default,
  pred_prob = predict(logit_fit_us, 
                      newdata = hold,
                      type = "prob")[, "Yes"])
hold_pred_us <- hold_pred_us %>%
  mutate(pred_class = factor(ifelse(hold_pred_us$pred_prob > 0.45, 
                                          "Yes", 
                                          "No")))
```


## Evaluate the Model
### Confusion Matrix

```{r}
confusionMatrix(hold_pred_us$pred_class, 
                hold_pred_us$Predicted_Loan_Default, 
                positive = "Yes")
```

### ROC Curve
The generation of the ROC curve is done via the two chunks below. Additionally, the corresponding AUC and F1 value are reported below. 

**Compute the ROC curve**
```{r}
preds_us <- ROCR::prediction(hold_pred_us$pred_prob, 
                             hold_pred_us$Predicted_Loan_Default)
perf_us <- ROCR::performance(preds_us, "tpr", "fpr")
```

**Plot Curve**
```{r}
plot(perf_us, col = "blue", lwd = 2, main = "ROC Curve fo rOur Model")
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)
```

**Extract AUC**
Here, we get the auc value. 
```{r}
auc_us <- ROCR::performance(preds_us, "auc")@y.values[[1]]
print(paste("The AUC is ",round(auc_us,2)))
```

### F1 Score
```{r}
F1_us <- MLmetrics::F1_Score(y_pred = hold_pred_us$pred_class,
                            y_true = hold$loan_default,
                            positive = "Yes")


print(paste("The F1 score is ",round(F1_us,2)))
```

# Conclusion
The undersampled and original datasets showed similar F1 scores. However, the sensitivity for the model trained using the original dataset. Thus, the better dataset would be considered to be that trained with the whole dataset because sensitivity reflects the ability of the model to capture the positive class, which in this case is significantly more influential than capturing the negative class. 

**New Imbalance Dataset**
| threshold | F1 Score |
|-----------|----------|
|  0.8     |  0.13    |
|  0.7     |  0.13    |
| 0.5      |  0.24        |
| 0.3      |  0.43        |
| 0.2      |  0.45        |
| 0.15      |    0.43      |

**New Imbalance Dataset**
| threshold | F1 Score |
|-----------|----------|
|  0.5     |  0.45    |
|  0.4      |  0.43    |
| 0.3      |  0.        |
| 0.6      |  0.45        |
| 0.7      |  0.44        |
| 0.8      |    0.43      |
