---
title: "Preprocessing the Data to Predict Loan Default"
author: "storera17"
output:
    html_document:
        code_folding: show
        df_print: paged
        number_sections: yes
        theme: readable
        toc: yes
        toc_float: yes
        code_download:
    word_document: yes
        toc: no
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE,warning=FALSE)
knitr::opts_chunk$set(fig.width=6, fig.height=4)

if(require(pacman)==0)
   {install.packages("pacman")}
pacman::p_load(devtools,caret,cluster,dplyr,fastDummies,leaps,pacman,tidyverse,skimr,fastDummies,GGally,DataExplorer,ggrepel,ggthemes,dslabs,lubridate,janitor,e1071,scales,recipes,magrittr,forcats,tidycensus,zipcodeR,tidygeocoder,tigris,sf, purrr,ROCR,rpart,rpart.plot, doParallel)

if (!require(mlba)) {
  library(devtools)
  install_github("gedeck/mlba/mlba", force=TRUE)
}
pacman::p_load(mlba,tidyverse)
```
# Data Overview & Summary Statistics

## Loading in the Data & Checking Data Type

> Training data set was coined `train.csv` and is not provided. 

```{r}
df_raw <- read_csv("train.csv", show_col_types = FALSE)

# Convert response variable to factor
df_raw <- df_raw %>%
  mutate(loan_default=factor(loan_default, levels=c("No","Yes")))

# Preview data
str(df_raw)
```

## Data Management

### Numeric Variables 

> The numeric variables were gathered together into `df_num` separate from the categorical variables for closer analysis. The variables were examined using the `str()` function to confirm all of the gathered variables were numeric.

```{r}
df_num <- df_raw %>%
  dplyr::select(where(is.numeric))
```

> The number of unqiue instances was captured using the following code below:

```{r}
sapply(df_num, n_unique)
```

### Converting Numeric Variables into Factors & Remove from 'df_num'

```{r}
# Convert to factors
df_raw$inq_last_6mths <- as.factor(df_raw$inq_last_6mths)
df_raw$acc_now_delinq <- as.factor(df_raw$acc_now_delinq)

# Remove them
df_num <- df_num %>%
  select(-inq_last_6mths,
         -acc_now_delinq)
```

### Categorical Variables

> The numeric variables were gathered together into `df_cat` separate from the numeric variables. The variables were examined using the `str()` function to confirm all of the gathered variables were categoricals.

```{r}
df_cat <- df_raw %>%
  dplyr::select(where(\(x) is.character(x) || is.factor(x)))
```

### Converting Character Variables to Date Variables

```{r}
df_cat <- df_cat %>%
  mutate(issue_d = lubridate::my(issue_d),
    earliest_cr_line = lubridate::my(earliest_cr_line),
    last_credit_pull_d = lubridate::my(last_credit_pull_d)
  )
```

> Next, we will examine the number of unique data values for each categorical variable. This will allow us to determine which variable should be converted into a factor.

```{r}
sapply(df_cat, n_unique)
```

### Converting Character Variables to factors

> Convert the variables mentioned above from characters to factors. Check if this was done successfully using the `str()` function.

```{r}
df_cat <- df_cat %>%
  mutate(term=factor(term, levels=c("36 months", "60 months")),
          initial_list_status=factor(initial_list_status, levels="f", "w")
  )
df_cat$emp_length <- as.factor(df_cat$emp_length)
df_cat$home_ownership <- as.factor(df_cat$home_ownership)
df_cat$verification_status <- as.factor(df_cat$verification_status)
df_cat$purpose <- as.factor(df_cat$purpose)
df_cat$initial_list_status <- as.factor(df_cat$initial_list_status)
df_cat$application_type <- as.factor(df_cat$application_type)
```

## Data Summaries

### Key Summary Statistics

> The key summary statistics are shown using the `skim()` function from the `skimr` package. The means of the numeric variables vary with the highest mean belonging to `annual_inc` at approximately 74,247, which is the average annual income, so it makes sense that a number would be this large for this variable. The smallest mean value of the numeric variables belonged to the variable `mort_acc` with a mean of about 1.8. The standard deviation of the numeric variables also had a wide range. The largest standard deviation also belonged to the variable `annual_inc` as well as the smallest standard deviation also belonged to `mort_acc`. The medians of the numeric variables also had a wide range with the smallest also being of the variable `mort_acc` and the largest belonging to `annual_inc`. Several variables had minimum values of zero which was the lowest minimum value. The largest maximum value belonged to the variable `annual_inc` at approximately 8,706,582.

```{r}
skimr::skim(df_num)
```

```{r}
skimr::skim(df_cat)
```

### Categorical Variables - Frequency Tables

```{r}
for (col in names(df)[sapply(df, is.factor)]) {
    print(table(df[[col]])[1:5])
}
```

# Data Visualization

## Visualiztion of Numeric Variable Distribution

### Histograms of the Numeric Variables

```{r}
DataExplorer::plot_histogram(df_num,
                             ggtheme = theme_minimal(),
                             ncol = 3,
                             nrow = 3)
```

### Box Plots of the Numeric Variables

> The boxplots are illustrations of the distributions and spread of data for the numeric variables across the loan default outcome.

```{r}
df_num_targ <- df_num %>%
  mutate(loan_default = df_raw$loan_default)
DataExplorer::plot_boxplot(
  df_num_targ,
  by = "loan_default",
  ggtheme = theme_minimal(),
  ncol = 3,
  nrow = 3)
```

### Visualization of Correlation between Variables with Heat Map

 > The strength and direction of linear relationships between numeric variables in the dataset is displayed in th heat map below. 

```{r}
DataExplorer::plot_correlation(df_num,
                               type="continuous",
                               cor_args = list("use" = "pairwise.complete.obs")
```

## Visualization of Categorical Variable Distribution with Bar characters

> The bar charts display the frequency distributions of the categorical variables in the dataset. The majority of loans for the response variable ‘loan_default’ are no, which means the majority of loans did not default during modeling

```{r}
# Select only categorical variables
DataExplorer::plot_bar(df_cat,
                       ggtheme = theme_minimal(),
                       ncol = 3,
                       nrow = 3)
```

# Handling Missing Data

## Tabular Summaries of Missing Values

### Missing Values in Numeric Values 

>  The numeric variables ‘revol_util’, ‘mort_acc’, and ‘mths_since_last_delinq’ contain missing values. The variable ‘mths_since_last_delinq’ is missing over 50% of data values. Thus, imputation will be required for this variable. The variable ‘mort_acc’ is missing a little less than 10% of data values, it will also required imputation.

```{r}
sapply(df_num, n_missing)
```

### Missing Values in Categorical Variables

> The categorical variables ‘emp_title’, ‘emp_length’, ‘title’, ‘pub_rec_bankruptcies’, and ‘last_credit_pull_d’ contain missing values. No categorical variable is misisng 10% or close to 10% of its values. Thus, all categorical variables will remain unchanged.

```{r}
sapply(df_cat, n_missing)
```

## Graphical illustrations of Missing Data

> The numeric and categorical variables were combined and the missing values were examined using the plot_missing() function. As stated previously, the variable ‘mths_since_last_delinq’ contains over 50% of missing values. In addition, the variable ‘mort_acc’ contains a little less than 10% of missing values.

```{r} 
df <- bind_cols(df_cat, df_num)
plot_missing(df)
```

## Simple Imputation: Fill Numerics with medians

> We decided to use a simple imputation for missing values. Ultimately, the method we chose was using the median in the place of missing values. We use the plot_missing() function to see if the missing data values are replace

```{r}
# Compute medians from TRAINING dataset
train_medians <- sapply(df, function(x)
  if (is.numeric(x)) median(x, na.rm = TRUE) else NA)

# Impute numeric variables using training medians
for (v in c("mort_acc", "pub_rec_bankruptcies", "revol_util")) {
  df[[v]][is.na(df[[v]])] <- train_medians[[v]]}
```

## Flag Method 

```{r}
# Flag for mths_since_last_delinq
df$mths_since_last_delinq[is.na(df$mths_since_last_delinq)] <- 999
df$no_delinq_flag <- ifelse(df$mths_since_last_delinq == 999, 1, 0) # Flag that preserves info that borrower has never been deliquent (according to survey) and lets model learn from it
```

## Use Missing Level 

```{r}
# 'initial_list_status'
df$initial_list_status <- forcats::fct_explicit_na(
  df$initial_list_status, na_level = "Missing")
# emp_length
df$emp_length <- forcats::fct_explicit_na(as.factor(df$emp_length), na_level = "Missing")
```

## Plot Missing to Check

```{r}
plot_missing(df)
```

# Outlier & Anomaly Detection

## Examination & Approach to Outliers

> We examined potential outliers using the interquartile range method. The values reported from the code below show the number of values that are larger than the interquartile range threshold (1.5 x IQR) for each numeric variable. These values do not necessarily represent errors or anomalies. These values are a reflection of variability within the data and are logical representations of real-world finanical data. Forthwidth, it would not be an accurate model of the data if we were to remove these high values.

```{r}
iqr_check <- function(x) {
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  sum(x < (q1 - 1.5 * iqr) | x > (q3 + 1.5 * iqr), na.rm = TRUE)
}

outlier_summary <- sapply(df_num, iqr_check)

# Print summary
outlier_summary
```

```{r}
# Evaluate
if (all(outlier_summary == 0)) {
  message("No significant outliers detected based on IQR method.")
}
```

# Data Transformation

## Transforming Highly Skewed Variables

> During the qualitative analysis of histograms for the numeric variables, we determined that ‘loan_amnt’, ‘installment’, ‘annual_inc’, ‘revol_bal’, ‘revol_util’, ‘dti’, and ‘mths_since_last_delinq’ would benefit from a logarithmic transformation because of being all right-skewed. This will tempor the skewed shaped of these variables and be more favorable for the model. After transforming the variables to a logarithmic scale, the skewness was reduced.

```{r}
skew_num <- c('annual_inc')

for (v in skew_num) {
  if (v %in% names(df)) {
    df[[paste0(v, "_log")]] <- log1p(df[[v]])
  }
}

skew_before <- sapply(df[skew_num], skewness, na.rm = TRUE)
skew_after <- sapply(df[paste0(skew_num, "_log")], skewness, na.rm = TRUE)

tibble::tibble(variable = skew_num,
               before_skew = round(skew_before, 2),
               after_skew = round(skew_after, 2))
```

# Dimension Reduction

## Correlation Examination

> The correlation heat map revealed a strong correlation between ‘loan_amnt’ with ‘installment’ and ‘fico_range_high’ with ‘fico_range_low’. In order to prevent multicolinearity, one of these variables needs to be removed from each pair. Here, we have just another way we computed a correlation matrix and found that the variables listed are highly correlated. This just further emphasizes what we explained when examining the correlation heat map previously. Start with removing the original non-transformed columns. Even removed variables are shown due to the ‘findCorrelation()’ function. This is demonstrated by previewing the first few rows.

```{r}
#remove <- c('annual_inc')
#df <- df %>%
#  select(-remove)

num_cols <- names(df)[sapply(df, is.numeric)]
cor_matrix <- cor(df[num_cols], use="pairwise.complete.obs")
high_corr <- findCorrelation(cor_matrix, cutoff = 0.85, names = TRUE, exact = TRUE)
print(high_corr)
```

## Removal of Redundancies / Irrelevant Features

```{r}
# Irrelevant Features
df <- df %>%
  select(-emp_title, 
         -title, 
#         -sub_grade, 
         -address, 
         -hardship_flag, 
         -debt_settlement_flag) #%>%
# Redundant (High correlation)
#  select(-installment)
```

# Feature Creation

## Binned Variables & Ordered Factors

### Binned Variables 

> The categorical variables ‘inq_last_6mths’, ‘acc_now_delinq’ need to be converted into categorical bins to capture risk tiers. These will be included in dummy encoding.

```{r}
df <- df %>%
  mutate(
    # Replace NA with 0 first
    acc_now_delinq = ifelse(is.na(acc_now_delinq), 0, acc_now_delinq),
    inq_last_6mths = ifelse(is.na(inq_last_6mths), 0, inq_last_6mths),

    # Create binned versions
    acc_now_delinq_binned = case_when(
      acc_now_delinq == 0 ~ "0",
      acc_now_delinq == 1 ~ "1",
      acc_now_delinq >= 2 ~ "2 or more"
    ),
    inq_last_6mths_binned = case_when(
      inq_last_6mths == 0 ~ "0",
      inq_last_6mths == 1 ~ "1",
      inq_last_6mths >= 2 ~ "2 or more"
    )
  )

# Convert to ordered factors
df$acc_now_delinq_binned <- factor(df$acc_now_delinq_binned,
                                   levels = c("0", "1", "2 or more"),
                                   ordered = TRUE)

df$inq_last_6mths_binned <- factor(df$inq_last_6mths_binned,
                                   levels = c("0", "1", "2 or more"),
                                   ordered = TRUE)
df <- df %>%
  select(-inq_last_6mths, -acc_now_delinq)
#df <- fastDummies::dummy_cols(df, select_columns = acc_now_delinq_binned, inq_last_6mths_binned,
#                              remove_first_dummy=TRUE)
```

> There is one distinct category that dominates. In order to capture the possible usefulness, we decided to combine the two minor categories. If no usefulnessa is provided we can drop this variable.

```{r}
# Convert back to character
df$application_type <- as.character(df$application_type)

# Combine low count variables
df$application_type[df$application_type %in% c("JOINT", "DIRECT_PAY")] <- "OTHER"

# Convert back to factor
df$application_type <- factor(df$application_type, levels = c("INDIVIDUAL", "OTHER"))
```

### Ordered Factors

```{r}
# Convert issue_d into issue_year and loan_age_days
df <- df %>%
  mutate(
    issue_year = lubridate::year(issue_d),
    loan_age_days = as.numeric(Sys.Date() - issue_d)
  ) %>%
  select(-issue_d) %>%
  mutate(
    issue_year = factor(issue_year,
                        levels = sort(unique(issue_year)),
                        ordered = TRUE)
  )
```

```{r}
# Define sub_grade order
sub_levels <- c("A1","A2","A3","A4","A5",
                "B1","B2","B3","B4","B5",
                "C1","C2","C3","C4","C5",
                "D1","D2","D3","D4","D5",
                "E1","E2","E3","E4","E5",
                "F1","F2","F3","F4","F5",
                "G1","G2","G3","G4","G5")

# Apply to hold
df$sub_grade <- factor(df$sub_grade, levels = sub_levels, ordered = TRUE)
df$grade <- substr(df$sub_grade, 1, 1)
df$grade <- factor(df$grade,
                     levels = c("A","B","C","D","E","F","G"),
                     ordered = TRUE)
df$sub_grade_num <- as.numeric(df$sub_grade)
```

## Ratio-Based Features

```{r}
df <- df %>%
  mutate(
    fico_score = (fico_range_low + fico_range_high) / 2
  ) %>%
  select(-fico_range_low, -fico_range_high)
```

```{r}
df <- df %>%
  # Engineer installment direct proportionality with annual_inc and loan_amnt
  mutate(installment_to_income = installment / annual_inc,
         installment_to_loanamt = installment / loan_amnt) %>%
  
  # Remove 'annual_inc' and 'installment'
  select(-installment, -annual_inc)

#df <- df %>%
#  mutate(
#    installment_to_income = log1p(installment_to_income),
#    installment_to_loanamt = log1p(installment_to_loanamt)) %>%
#  select(-installment_to_income, -installment_to_loanamt)
```

## Time-Based / Duration Variables

```{r}
# Convert 'earliest_cr_line' and 'last_credit_pull_d' into a numeric interval. This represents borrower history length. 
df <- df %>%
  mutate(
    earliest_cr_line = as.Date(earliest_cr_line, format = "%Y-%m-%d"),
    last_credit_pull_d = as.Date(last_credit_pull_d, format = "%Y-%m-%d"),
    credit_history_days = as.numeric(difftime(last_credit_pull_d, earliest_cr_line, units = "days"))) %>%
# Remove original components
  select(-last_credit_pull_d, -earliest_cr_line)
```

# Final Checks

## Count Missing

```{r}
colSums(is.na(df))
```

> We located missing data in the variable ‘credit_history_days’ and decided to impute it.

```{r}
# compute median from training dataset
training_credit_median <- median(df$credit_history_days, na.rm = TRUE)

# apply training median to training set
df <- df %>%
  mutate(
    credit_history_days = ifelse(
      is.na(credit_history_days),
      training_credit_median,
      credit_history_days))
```

## Confirm No Missing Data

```{r}
colSums(is.na(df))
```

```{r}
# Examine dataset
str(df)
```

```{r}
skimr::skim(df)
```

## Save Data - RDS File 

```{r}
saveRDS(df, file = 'storera17_train.rds')
```

# Holdout Preprocessing

## Initials 

```{r}
# Load data
hold_raw <- read_csv("holdout.csv", show_col_types = FALSE)

# Change response variable to classification
hold_raw <- hold_raw %>%
  mutate(loan_default=factor(loan_default, levels=c("No","Yes")))

# Preview data
str(hold_raw)
```

## Numeric Processing 

```{r}
# Select only numeric variables
hold_num <- hold_raw %>%
  dplyr::select(where(is.numeric))

# Convert numeric variables into factors
hold_raw$inq_last_6mths <- as.factor(hold_raw$inq_last_6mths)
hold_raw$acc_now_delinq <- as.factor(hold_raw$acc_now_delinq)

# Remove non-numeric variables form numerics
hold_num <- hold_num %>%
  select(-inq_last_6mths,
         -acc_now_delinq)
```

## Categorical Processing

```{r}
# Select only categorical variables
hold_cat <- hold_raw %>%
  dplyr::select(where(\(x) is.character(x) || is.factor(x)))

# Convert applicable character variables to dates
hold_cat <- hold_cat %>%
  mutate(issue_d = lubridate::my(issue_d),
    earliest_cr_line = lubridate::my(earliest_cr_line),
    last_credit_pull_d = lubridate::my(last_credit_pull_d)
  )

# Convert character variables to factors
hold_cat <- hold_cat %>%
  mutate(term=factor(term, levels=c("36 months", "60 months")),
          initial_list_status=factor(initial_list_status, levels="f", "w"))
hold_cat$emp_length <- as.factor(hold_cat$emp_length)
hold_cat$home_ownership <- as.factor(hold_cat$home_ownership)
hold_cat$verification_status <- as.factor(hold_cat$verification_status)
hold_cat$purpose <- as.factor(hold_cat$purpose)
hold_cat$initial_list_status <- as.factor(hold_cat$initial_list_status)
hold_cat$application_type <- as.factor(hold_cat$application_type)
```

## Data Summary

```{r} 
# Data summary of numerics
skimr::skim(hold_num)

# Data summary of categoricals
skimr::skim(hold_cat)
```

## Data Visualization

```{r}
# Numeric frequency figures (histograms)
DataExplorer::plot_histogram(hold_num,
                             ggtheme = theme_minimal(),
                             ncol = 3,
                             nrow = 3)

# Numeric boxplots
hold_num_targ <- hold_num %>%
  mutate(loan_default = hold_raw$loan_default)
DataExplorer::plot_boxplot(
  hold_num_targ,
  by = "loan_default",
  ggtheme = theme_minimal(),
  ncol = 3,
  nrow = 3)

# Numeric correlation using heat map
DataExplorer::plot_correlation(hold_num,
                               type="continuous",
                               cor_args = list("use" = "pairwise.complete.obs")
                              )

# Categorical bar charts
DataExplorer::plot_bar(hold_cat,
                       ggtheme = theme_minimal(),
                       ncol = 3,
                       nrow = 3)
```

## Combine Numerics with categoricals

```{r}
hold <- bind_cols(hold_cat, hold_num)
```

# Handling Missing Data

## Evaluation of Missingness

```{r}
# Tabular missing data summary
colSums(is.na(hold))

# Graphical missing data summary
plot_missing(hold)
```

## Imputation

### Median Imputation Method

```{r}
# perform variable imputation
for (v in c("mort_acc", "pub_rec_bankruptcies", "revol_util")) {
  # Holdout data (use training medians)
  hold[[v]][is.na(hold[[v]])] <- train_medians[[v]]
}
```

### Flag Method

```{r}
# Flag for mths_since_last_delinq
hold$mths_since_last_delinq[is.na(hold$mths_since_last_delinq)] <- 999
hold$no_delinq_flag <- ifelse(hold$mths_since_last_delinq == 999, 1, 0) # Flag that preserves info that borrower has never been deliquent (according to survey) and lets model learn from it
```

### Use Missing Level 

```{r}
# 'initial_list_status'
hold$initial_list_status <- forcats::fct_explicit_na(
  hold$initial_list_status, na_level = "Missing")
# emp_length
hold$emp_length <- forcats::fct_explicit_na(as.factor(hold$emp_length), na_level = "Missing")
```

### Plot Missing to Check

```{r}
plot_missing(hold)
```

## Outlier & Anomaly Detection

```{r}
hold_iqr_check <- function(x) {
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  sum(x < (q1 - 1.5 * iqr) | x > (q3 + 1.5 * iqr), na.rm = TRUE)
}

hold_outlier_summary <- sapply(hold_num, hold_iqr_check)

# Print summary
hold_outlier_summary

# Evaluate
if (all(hold_outlier_summary == 0)) {
  message("No significant outliers detected based on IQR method.")
}
```

## Data Transfotmation

```{r}
hold_skew_num <- c('annual_inc')

for (v in hold_skew_num) {
  if (v %in% names(hold)) {
    hold[[paste0(v, "_log")]] <- log1p(hold[[v]])
  }
}

hold_skew_before <- sapply(hold[hold_skew_num], skewness, na.rm = TRUE)
hold_skew_after <- sapply(hold[paste0(hold_skew_num, "_log")], skewness, na.rm = TRUE)

tibble::tibble(variable = hold_skew_num,
               hold_before_skew = round(hold_skew_before, 2),
               hold_after_skew = round(hold_skew_after, 2))
```

## Dimension Reduction

```{r}
#hold_remove <- c('annual_inc')
#hold <- hold %>%
#  select(-hold_remove)

hold_num_cols <- names(hold)[sapply(hold, is.numeric)]
hold_cor_matrix <- cor(hold[hold_num_cols], use="pairwise.complete.obs")
hold_high_corr <- findCorrelation(hold_cor_matrix, cutoff = 0.85, names = TRUE, exact = TRUE)
print(hold_high_corr)
```

## Removal of Redundancie / Irrelevant Features


```{r}
# Irrelevant Features
hold <- hold %>%
  select(-emp_title,
         -title, 
#         -sub_grade, 
         -address, 
         -hardship_flag, 
         -debt_settlement_flag)  #%>%
# Redundant (High correlation)
#  select(-installment)
```

## Feature Creation

### Binned variables and ordered factors
```{r}
hold <- hold %>%
  mutate(
    # Replace NA with 0 first
    acc_now_delinq = ifelse(is.na(acc_now_delinq), 0, acc_now_delinq),
    inq_last_6mths = ifelse(is.na(inq_last_6mths), 0, inq_last_6mths),

    # Create binned versions
    acc_now_delinq_binned = case_when(
      acc_now_delinq == 0 ~ "0",
      acc_now_delinq == 1 ~ "1",
      acc_now_delinq >= 2 ~ "2 or more"
    ),
    inq_last_6mths_binned = case_when(
      inq_last_6mths == 0 ~ "0",
      inq_last_6mths == 1 ~ "1",
      inq_last_6mths >= 2 ~ "2 or more"
    )
  )

# Convert to ordered factors
hold$acc_now_delinq_binned <- factor(hold$acc_now_delinq_binned,
                                   levels = c("0", "1", "2 or more"),
                                   ordered = TRUE)

hold$inq_last_6mths_binned <- factor(hold$inq_last_6mths_binned,
                                   levels = c("0", "1", "2 or more"),
                                   ordered = TRUE)
# Remove original variables
hold <- hold %>%
  select(-inq_last_6mths, -acc_now_delinq)

# Convert back to character
hold$application_type <- as.character(hold$application_type)

# Combine low count variables
hold$application_type[hold$application_type %in% c("JOINT", "DIRECT_PAY")] <- "OTHER"

# Convert back to factor
hold$application_type <- factor(hold$application_type, levels = c("INDIVIDUAL", "OTHER"))

hold$home_ownership <- fct_collapse(hold$home_ownership,
                                        OTHER = c("OTHER", "ANY"))

# Convert issue_d into issue_year and loan_age_days
hold <- hold %>%
  mutate(
    issue_year = lubridate::year(issue_d),
    loan_age_days = as.numeric(Sys.Date() - issue_d)
  ) %>%
  select(-issue_d) %>%
  mutate(
    issue_year = factor(issue_year,
                        levels = sort(unique(issue_year)),
                        ordered = TRUE)
  )

# Define sub_grade order
sub_levels <- c("A1","A2","A3","A4","A5",
                "B1","B2","B3","B4","B5",
                "C1","C2","C3","C4","C5",
                "D1","D2","D3","D4","D5",
                "E1","E2","E3","E4","E5",
                "F1","F2","F3","F4","F5",
                "G1","G2","G3","G4","G5")

# Apply to hold
hold$sub_grade <- factor(hold$sub_grade, levels = sub_levels, ordered = TRUE)
hold$grade <- substr(hold$sub_grade, 1, 1)
hold$grade <- factor(hold$grade,
                     levels = c("A","B","C","D","E","F","G"),
                     ordered = TRUE)
hold$sub_grade_num <- as.numeric(hold$sub_grade)
```

### Ratio-Based Features

```{r}
hold <- hold %>%
  mutate(
    fico_score = (fico_range_low + fico_range_high) / 2
  ) %>%
  select(-fico_range_low, -fico_range_high)

hold <- hold %>%
  # Engineer installment direct proportionality with annual_inc and loan_amnt
  mutate(installment_to_income = installment / annual_inc,
         installment_to_loanamt = installment / loan_amnt) %>%
  
  # Remove 'annual_inc' and 'installment'
  select(-installment, -annual_inc)

#hold <- hold %>%
#  mutate(
#    installment_to_income = log1p(installment_to_income),
#    installment_to_loanamt = log1p(installment_to_loanamt)) %>%
#  select(-installment_to_income, -installment_to_loanamt)
```

### Time-Based / Duration Variables

```{r}
# Convert 'earliest_cr_line' and 'last_credit_pull_d' into a numeric interval. This represents borrower history length. 
hold <- hold %>%
  mutate(
    earliest_cr_line = as.Date(earliest_cr_line, format = "%Y-%m-%d"),
    last_credit_pull_d = as.Date(last_credit_pull_d, format = "%Y-%m-%d"),
    credit_history_days = as.numeric(difftime(last_credit_pull_d, earliest_cr_line, units = "days"))) %>%
# Remove original components
  select(-last_credit_pull_d, -earliest_cr_line)
```

## Last Checks

### Count Missing

```{r}
colSums(is.na(hold))
```

> Impute for missing values (use training imputation)

```{r}
hold <- hold %>%
  mutate(
    credit_history_days = ifelse(
      is.na(credit_history_days),
      training_credit_median,
      credit_history_days))
```

```{r}
# Confirm no missing data
colSums(is.na(hold))

# Examine dataset
str(hold)

skimr::skim(hold)
```

## Save Data

```{r}
saveRDS(hold, file = 'storera17_holdout.rds')
```