---
title: "Preprocessing the Data to Predict Loan Default"
author: "storera17"
output:
    html_document:
        code_folding: show
        df_print: paged
        number_sections: yes
        theme: readable
        toc: yes
        toc_float: yes
        code_download:
    word_document: yes
        toc: no
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE,warning=FALSE)
knitr::opts_chunk$set(fig.width=6, fig.height=4)

if(require(pacman)==0)
   {install.packages("pacman")}
pacman::p_load(devtools,caret,cluster,dplyr,fastDummies,leaps,pacman,tidyverse,skimr,fastDummies,GGally,DataExplorer,ggrepel,ggthemes,dslabs,lubridate,janitor,e1071,scales,recipes,magrittr,forcats,tidycensus,zipcodeR,tidygeocoder,tigris,sf, purrr,ROCR,rpart,rpart.plot, doParallel)

if (!require(mlba)) {
  library(devtools)
  install_github("gedeck/mlba/mlba", force=TRUE)
}
pacman::p_load(mlba,tidyverse)
```
# Data Overview & Summary Statistics

## Loading in the Data & Checking Data Type

> Training data set was coined `train.csv` and is not provided. 

```{r}
df_raw <- read_csv("train.csv", show_col_types = FALSE)

# Convert response variable to factor
df_raw <- df_raw %>%
  mutate(loan_default=factor(loan_default, levels=c("No","Yes")))

# Preview data
str(df_raw)
```

## Data Management

### Numeric Variables 

> The numeric variables were gathered together into `df_num` separate from the categorical variables for closer analysis. The variables were examined using the `str()` function to confirm all of the gathered variables were numeric.

```{r}
df_num <- df_raw %>%
  dplyr::select(where(is.numeric))
```

> The number of unqiue instances was captured using the following code below:

```{r}
sapply(df_num, n_unique)
```

### Converting Numeric Variables into Factors & Remove from 'df_num'

```{r}
# Convert to factors
df_raw$inq_last_6mths <- as.factor(df_raw$inq_last_6mths)
df_raw$acc_now_delinq <- as.factor(df_raw$acc_now_delinq)

# Remove them
df_num <- df_num %>%
  select(-inq_last_6mths,
         -acc_now_delinq)
```

### Categorical Variables

> The numeric variables were gathered together into `df_cat` separate from the numeric variables. The variables were examined using the `str()` function to confirm all of the gathered variables were categoricals.

```{r}
df_cat <- df_raw %>%
  dplyr::select(where(\(x) is.character(x) || is.factor(x)))
```

### Converting Character Variables to Date Variables

```{r}
df_cat <- df_cat %>%
  mutate(issue_d = lubridate::my(issue_d),
    earliest_cr_line = lubridate::my(earliest_cr_line),
    last_credit_pull_d = lubridate::my(last_credit_pull_d)
  )
```

> Next, we will examine the number of unique data values for each categorical variable. This will allow us to determine which variable should be converted into a factor.

```{r}
sapply(df_cat, n_unique)
```

### Converting Character Variables to factors

> Convert the variables mentioned above from characters to factors. Check if this was done successfully using the `str()` function.

```{r}
df_cat <- df_cat %>%
  mutate(term=factor(term, levels=c("36 months", "60 months")),
          initial_list_status=factor(initial_list_status, levels="f", "w")
  )
df_cat$emp_length <- as.factor(df_cat$emp_length)
df_cat$home_ownership <- as.factor(df_cat$home_ownership)
df_cat$verification_status <- as.factor(df_cat$verification_status)
df_cat$purpose <- as.factor(df_cat$purpose)
df_cat$initial_list_status <- as.factor(df_cat$initial_list_status)
df_cat$application_type <- as.factor(df_cat$application_type)
```

## Data Summaries

### Key Summary Statistics

> The key summary statistics are shown using the `skim()` function from the `skimr` package. The means of the numeric variables vary with the highest mean belonging to `annual_inc` at approximately 74,247, which is the average annual income, so it makes sense that a number would be this large for this variable. The smallest mean value of the numeric variables belonged to the variable `mort_acc` with a mean of about 1.8. The standard deviation of the numeric variables also had a wide range. The largest standard deviation also belonged to the variable `annual_inc` as well as the smallest standard deviation also belonged to `mort_acc`. The medians of the numeric variables also had a wide range with the smallest also being of the variable `mort_acc` and the largest belonging to `annual_inc`. Several variables had minimum values of zero which was the lowest minimum value. The largest maximum value belonged to the variable `annual_inc` at approximately 8,706,582.

```{r}
skimr::skim(df_num)
```

```{r}
skimr::skim(df_cat)
```

### Categorical Variables - Frequency Tables

```{r}
for (col in names(df)[sapply(df, is.factor)]) {
    print(table(df[[col]])[1:5])
}
```

# Data Visualization

## Visualiztion of Numeric Variable Distribution

### Histograms of the NUmeric Variables

```{r}
DataExplorer::plot_histogram(df_num,
                             ggtheme = theme_minimal(),
                             ncol = 3,
                             nrow = 3)
```


